---
title: "North Campus Open Space bird community analysis: survey years 1-8 (2017-2025) "
format: pdf
execute: 
  echo: false
fig-align: center
author: Matt Vinh
---

```{r setup}
#| include: false

library(here)
library(tidyverse)
library(vegan)
library(INLA)
library(knitr)
library(DT)
library(ggrepel)
library(showtext)

datapath <- here("data","Matt_Vinh","compiled_and_cleaned_2026-02-18.rds")
data <- read_rds(datapath)

surveypath <- here("data","Matt_Vinh","survey-level_2026-03-01.rds")
survey <- read_rds(surveypath)

SBApath <- here("data","Matt_Vinh","SBA_hourly_precipitation_2026-03-01.rds")
precip <- read_rds(SBApath)

# This rds is not synced to github... you have to download it from Google Drive
#It includes all ebird checklists for the NCOS hotspot... more filtering required
datapath_ebird <- here("data","ncos_data.rds")
ncos_ebird <- read_rds(datapath_ebird) %>% 
  mutate(observation_count = na_if(observation_count,"X"),
         observation_count = as.integer(observation_count))

```

## Abstract (Executive summary)

The short-term effects of the recent North Campus Open Space (NCOS) restoration project have been individually observed as massive, beneficial changes to the various ecosystems of the area. However, such changes lack a formal analysis on a higher level than the annual reports of statistics. To explore and understand how the restoration has affected NCOS avian biodiversity, the bird survey data of the Cheadle Center for Biodiversity and Restoration over eight years was leveraged.

The division of avian biodiversity into breeding versus non breeding species observed at NCOS provides key insight into the effect of the restoration on bird  populations - differentiating the true dependent species as opposed to those who merely visit NCOS temporarily can assist detected trend explanation. Additionally, accounting for the short duration of available data is crucial for correct interpretation of statistical findings, as trends become harder to detect and less representative with fewer data.

## Introduction

## Methods

### Data collection

one expert birder per survey vs two per team per survey?

within one hour after sunrise vs 45 mins after sunrise?

Starting in September 2017, following the completion of the restoration of North Campus Open Space (NCOS), Cheadle Center staff and associates conduct monthly bird surveys along two trails that cover the entirety of NCOS. The routes begin at Venocco Bridge and diverge west and east until the two groups meet at the survey end near Phelps Creek on the northern side of NCOS. The survey begins (within one hour vs 45 mins after) sunrise and takes 2 to 2.5 hours. Using binoculars, spotting scopes and with the assistance of (1 vs 4) expert birders, all birds observed or heard are documented on the GIS app Field Maps with their species and abundance. The Field Maps app also records the obeserver's location every 30 seconds. Observations that are considered repeats, whether by the same group or due to overlap of observation between the two groups, are marked as repeats. At the conclusion of the survey, the expert birders of the survey verify the collected bird observations and submit the data to the Cornell Lab of Ornithology's eBird repository, excluding repeat observations. Over 8 years, CCBER has conducted 96 surveys, recording 15,644 observations of 173 (vs ncos_ebird 258) different species.

```{r}

number_species_ccber <- data %>% filter(category == "species")
length(unique(number_species_ccber$species))

number_species_ebird <- ncos_ebird %>% filter(category == "species")
length(unique(number_species_ebird$common_name))

```

### Data cleaning

-   summarize what happens in cleaning script

-   standardize species names according to XYZ (specific ebird taxonomy version)



### Data analysis

Species richness

Species composition (NMDS & guild trends)

Temporal trends for specific groups

-waterfowl, breeding birds (April-July, evidence of breeding in SB BBS), raptors...

Basic comparison to ebird data over same period

## Results

### Data set overview

## Species richness over time

## Comparison to ebird data

```{r}
#| label: ebird comparison
#| message: false
#| warning: false

species_ebird <- ncos_ebird %>% 
  filter(category == "species") %>% 
  filter(observation_date > "2017-09-25") %>% 
  pull(common_name) %>% 
  unique()

species_monthly_surveys <- data %>% 
  filter(category == "species") %>% 
  pull(species) %>% 
  unique()

#what species are in ebird that are NOT in monthly surveys
setdiff(species_ebird, species_monthly_surveys)

#what species are in our monthly surveys but not in ebird (probably nothing!)
setdiff(species_monthly_surveys, species_ebird)

#one species! common gull. Maybe mis-identified?
# - after fixing of common gull species name, there are no species recorded
# - by ccber that are not also recorded by ebird



# understanding frequency of birds recorded on ebird but not ccber

diff <- setdiff(species_ebird, species_monthly_surveys)
difference_df <- data.frame(species = diff,total_count = length(diff) * NA)
restricted <- ncos_ebird %>% 
  filter(observation_date >= as.Date("2017-09-01") & observation_date <= as.Date("2025-08-31"))
for (i in 1:length(diff)){
  temp <- restricted %>% filter(common_name == diff[i])
  difference_df$total_count[i] = sum(temp$observation_count,na.rm = T)
}
print(difference_df)

```

## Species composition over time

```{r line plot}

# data organization

richness <- data %>% 
  filter(!survey_year == 9, category == "species") %>% 
  group_by(survey_year) %>% 
  summarise(n = length(unique(species)))

ebird_sy <- ncos_ebird %>% # sy = survey year
  mutate(survey_year = case_when(
      observation_date >= as.Date("2017-09-01") & observation_date <= as.Date("2018-08-31") ~ 1,
      observation_date >= as.Date("2018-09-01") & observation_date <= as.Date("2019-08-31") ~ 2,
      observation_date >= as.Date("2019-09-01") & observation_date <= as.Date("2020-08-31") ~ 3,
      observation_date >= as.Date("2020-09-01") & observation_date <= as.Date("2021-08-31") ~ 4,
      observation_date >= as.Date("2021-09-01") & observation_date <= as.Date("2022-08-31") ~ 5,
      observation_date >= as.Date("2022-09-01") & observation_date <= as.Date("2023-08-31") ~ 6,
      observation_date >= as.Date("2023-09-01") & observation_date <= as.Date("2024-08-31") ~ 7,
      observation_date >= as.Date("2024-09-01") & observation_date <= as.Date("2025-08-31") ~ 8,
      .default = NA))
ebird_richness <- ebird_sy %>% 
  filter(category == "species",
         !is.na(survey_year)) %>% 
  group_by(survey_year) %>% 
  summarize(richness = length(unique(common_name)))

richness_composite <- inner_join(richness,ebird_richness) %>% 
  mutate(diff = richness - n)

reports_path <- here("data","species_totals_from_ncos_annual_reports.csv")
annual_reports <- read_csv(reports_path)

# visualization

lines <- ggplot() + 
  geom_point(data = richness,
             aes(x = survey_year,
                 y = n),
             color = "blue",
             size = 3) + 
  geom_line(data = richness,
             aes(x = survey_year,
                 y = n),
             color = "blue",
            linewidth = 1.5) + 
  annotate("text",
           x = 7,
           y = richness$n[richness$survey_year == 7] - 11,
           label = "CCBER",
           color = "blue",
           size = 7) + 
  geom_point(data = ebird_richness,
             aes(x = survey_year,
                 y = richness),
             color = "red",
             size = 3) + 
  geom_line(data = ebird_richness,
             aes(x = survey_year,
                 y = richness),
             color = "red",
            linetype = "dashed",
            linewidth = 1.5) + 
  annotate("text",
           x = 7,
           y = ebird_richness$richness[ebird_richness$survey_year == 7] - 11,
           label = "eBird",
           color = "red",
           size = 7) + 
  geom_point(data = richness_composite,
             aes(x = survey_year,
                 y = diff),
             color = "purple",
             size = 3) + 
   geom_line(data = richness_composite,
             aes(x = survey_year,
                 y = diff),
             color = "purple",
             linetype = "longdash",
             linewidth = 1.5) + 
  annotate("text",
           x = 7,
           y = richness_composite$diff[richness_composite$survey_year == 7] - 11,
           label = "eBird - CCBER",
           color = "purple",
           size = 7) + 
  geom_point(data = annual_reports,
             aes(x = survey_year,
                 y = reported_richness),
             color = "forestgreen",
             size = 3) + 
  geom_line(data = annual_reports,
             aes(x = survey_year,
                 y = reported_richness),
             color = "forestgreen",
            linetype = "twodash",
            linewidth = 1.5) + 
  annotate("text",
           x = 6,
           y = annual_reports$reported_richness[annual_reports$survey_year == 7] - 11,
           label = "CCBER Annual Reports",
           color = "forestgreen",
           size = 7) + 
  theme_classic() + 
  labs(title = "Species Richness by Survey Year",
       x = "Survey Year",
       y = "Number of Species Detected") + 
  theme(text = element_text(family = "")
    plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 24),
        axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
        legend.position = "none") + 
  scale_x_continuous(breaks = seq(1,8,by = 1)) + 
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,225, by = 50),
                     limits = c(0,225))

temp1 <- richness %>% 
  rename(richness = n) %>% 
  mutate(category = "CCBER")
temp2 <- ebird_richness %>% 
  mutate(category = "eBird")
temp3 <- richness_composite %>% 
  select(c(1,4)) %>% 
  rename(richness = diff) %>% 
  mutate(category = "eBird_minus_CCBER")
temp4 <- annual_reports %>% 
  rename(richness = reported_richness) %>% 
  mutate(category = "CCBER_Annual_Reports") %>% 
  add_row(survey_year = 8, richness = 0, category = "CCBER_Annual_Reports")
richness_joined <- rbind(rbind(rbind(temp1, temp2), temp3), temp4)

ggplot(richness_joined,
       aes(x = survey_year,
           y = richness,
           fill = category)) + 
  geom_col(alpha = 0.8) + 
  facet_wrap(~ category) + 
  theme_classic() + 
  labs(title = "Species Richness by Survey Year",
       x = "Survey Year",
       y = "Number of Species Detected") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 14),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        legend.position = "none") + 
  scale_fill_manual(values = c("CCBER" = "blue",
                               "eBird" = "red",
                               "eBird_minus_CCBER" = "purple",
                               "CCBER_Annual_Reports" = "forestgreen"))

# saving

ggsave(filename = paste0("NCOS_Species_Richness_Comparison_",
                         format(Sys.Date(), "%y-%m-%d"),
                         ".svg"),
       lines,
       dpi = 1000,
       width = 9,
       height = 6,
       units = "in")

```

```{r line plot for poster}

lines2 <- ggplot() + 
  geom_point(data = richness,
             aes(x = survey_year,
                 y = n),
             color = "darkgreen",
             size = 3) + 
  geom_line(data = richness,
             aes(x = survey_year,
                 y = n),
             color = "darkgreen",
            linewidth = 1.5) + 
  annotate("text",
           x = 7,
           y = richness$n[richness$survey_year == 7] - 25,
           label = "Cheadle\nCenter",
           color = "darkgreen",
           size = 7) + 
  geom_point(data = ebird_richness,
             aes(x = survey_year,
                 y = richness),
             color = "saddlebrown",
             size = 3) + 
  geom_line(data = ebird_richness,
             aes(x = survey_year,
                 y = richness),
             color = "saddlebrown",
            linewidth = 1.5) + 
  annotate("text",
           x = 7,
           y = ebird_richness$richness[ebird_richness$survey_year == 7] - 15,
           label = "eBird",
           color = "saddlebrown",
           size = 7) + 
  theme_classic() + 
  labs(x = "Survey Year",
       y = "Number of Species\nDetected") + 
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 16),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 16),
        legend.position = "none",
        text = element_text(family = "Avenir LT Std")) + 
  ylim(c(75, 200)) + 
  scale_x_continuous(breaks = seq(1, 8, by = 1))

ggsave(filename = paste0("NCOS_Species_Richness_Comparison_poster_",
                         format(Sys.Date(), "%y-%m-%d"),
                         ".svg"),
       lines2,
       dpi = 1000,
       width = 9,
       height = 4,
       units = "in")

```

```{r max count nmds}
#| label: fig-nmds
#| message: false
#| warning: false
#| fig-cap: Revised NMDS based on 'Maximum Species Abundance' of each species in each year. I reproduced it thinking that with cleaner data it might appear differently, giving insight into species identity similarities or differences, following the bar chart of species richness. The usage of setdiff() may be more desirable in this situation.
#| include: false

year_max <- data %>% 
  filter(!survey_year == 9) %>%
  group_by(survey_year, species) %>%
  summarise(max_count = max(count))

community <- as.matrix(unclass(xtabs(max_count ~ survey_year + species,
                                     data = year_max)))

nmds <- metaMDS(community,autotransform = F,distance = "bray",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

nmds_site_scores <- as.data.frame(scores(nmds)$sites) %>% 
  rownames_to_column(var = "survey_year") %>% 
  mutate(survey_year = as.numeric(survey_year)) %>% 
  arrange(survey_year) %>% 
  mutate(
    xend = lead(NMDS1),
    yend = lead(NMDS2))

ggplot() + 
  geom_point(data = nmds_site_scores,
                      aes(x = NMDS1,
                          y = NMDS2,
                          color = survey_year),
             size = 3) + 
  geom_segment(data = nmds_site_scores,
               aes(x = NMDS1,
                   y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  annotate("text",
           x = -0.2,
           y = 0.15,
           label = paste("2D stress =",
                         round(nmds$stress,3))) + 
  scale_color_viridis_c(option = "C",
                        alpha = 0.8,
                        end = 0.95) + 
  theme_light() + 
  labs(title = "Maximum Species Abundance by Survey Year NMDS",
       color = "Survey Year") + 
  theme(plot.title = element_text(hjust = 0.5))

```



```{r jaccard binary nmds}
#| include: false
#| # TODO- this chunk was a rough attempt at creating an NMDS based on presence-absence (really presence vs. non-detection) of species rather than abundance.

nmds_binary <- metaMDS(community,autotransform = T, distance = "jaccard",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

nmds_site_scores_binary <- as.data.frame(scores(nmds_binary)$sites) %>% 
  rownames_to_column(var = "survey_year") %>% 
  mutate(survey_year = as.numeric(survey_year)) %>% 
  arrange(survey_year) %>% 
  mutate(
    xend = lead(NMDS1),
    yend = lead(NMDS2))

ggplot() + 
  geom_point(data = nmds_site_scores_binary,
                      aes(x = NMDS1,
                          y = NMDS2,
                          color = survey_year),
             size = 3) + 
  geom_segment(data = nmds_site_scores_binary,
               aes(x = NMDS1,
                   y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  annotate("text",
           x = -0.2,
           y = 0.15,
           label = paste("2D stress =",
                         round(nmds_binary$stress,3))) + 
  scale_color_viridis_c(option = "C",
                        alpha = 0.8,
                        end = 0.95) + 
  theme_light() + 
  labs(title = "NMDS based on Jaccard distance (presence/absence)",
       color = "Survey Year") + 
  theme(plot.title = element_text(hjust = 0.5))
```


```{r wright-ueda et al paper methods replication}

##### copy-paste template for note-taking ######################################

species <- 'CASJ'
U <- 0.007 # component of prior
# larger makes trend for both ggplots more apparent
# smaller makes trend of first ggplot undetectable, less prominent for second ggplot
distribution <- 'nbinomial' # distribution either poisson or nbinomial
# only use nbinomial when poisson is overdispersed

speciescol <- paste0('n.',species)
MyData <- na.exclude(df[, c(speciescol, # specifying variables to include
                            "Census_Area", "Day_of_year", "Date", 'Year',
                            'ID',
                            "Observer1")])
names(MyData)[1] <- 'Abundance'
bspecies <- MyData[,1] # only speciescol variable

hyper.prec <- list(theta = list(prior = "pc.prec", param = c(U, 0.05))) # prior
### hyper precision \ hyper parameters control prior
# pc.prec: penalized complexity priors, helps avoid overdispersion or over-fitting
# specifically, U and 0.05: P(sigma > 0.007) = 0.05 represents complexity criteria
# U range: 0.000125 to 1
# only expects random walk standard deviation to exceed 0.007 at 5% frequency
inla.setOption(scale.model.default = TRUE) #?

I1 <- inla(Abundance ~ f(Year, model = "rw1", hyper = hyper.prec) + # first order walk
             f(Day_of_year, model = "rw2", hyper = hyper.prec) + # second order walk
             # variables with suspected data correlation are assigned to random effects
             f(Census_Area, model = "iid") + # random effect
             f(Observer1, model = "iid"), # random effect
           # variables that impact temporal autocorrelation, where t is affected by t - 1,
           # are assigned to random walks
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           ### dic: deviance information criterion
           # - quantifies predictive fit, penalizes model complexity
           # - adjacent to AIC but for bayesian statistics
           ### waic: watanabe-akaike or widely-applicable information criterion
           # -  balances model fit and complexity, estimates out-of-sample predictive performance
           family = distribution,
           data = MyData)

### "All models employed weakly informative penalized complexity priors"
# - pc.prec, hyper parameters and information criteria (dic, waic) agree with summary

################################################################################

survey_path <- here("data","aggregated","Matt_Vinh","survey-level_2026-02-18.rds")
df <- read_rds(survey_path)

species <- 'Killdeer'
U <- 0.05
distribution <- 'poisson' # poisson seems better than nbinomial, however have not tested overdispersion

MyData <- na.exclude(df[, c(species,
                            "survey_id",
                            "day_of_survey_year",
                            "observation_date",
                            "survey_year")])
names(MyData)[1] <- 'Abundance'
bspecies <- MyData[,1]

hyper.prec <- list(theta = list(prior = "pc.prec", param = c(U, 0.05)))
inla.setOption(scale.model.default = TRUE)

I1 <- inla(Abundance ~ f(survey_year, model = "rw1", hyper = hyper.prec) + 
             f(day_of_survey_year, model = "rw2", hyper = hyper.prec),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)


#### Model Validation ####
MyData$Fitted <- I1$summary.fitted.values[,'mean']
# Plot trend for year
Yearsm <- I1$summary.random$survey_year
ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = Yearsm, aes(x = ID, y = mean)) +
  geom_ribbon(data = Yearsm, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  ggtitle('Trend by Year') +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(Yearsm$`0.025quant`)-0.05,max(Yearsm$`0.975quant`)+0.05)


### Dispersion ###
mu <- I1$summary.fitted.values[, "mean"]
theta <- I1$summary.hyperpar[1, "mean"]

if(distribution == 'nbinomial'){
  # Pearson residuals for negative binomial distribution
  E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
}
if(distribution == 'poisson'){
  # Pearson residuals for poisson distribution
  E1 <- (bspecies - mu) / sqrt(mu) 
}
if(distribution == 'gaussian'){
  # Residuals for gaussian distribution
  ### there is exactly one species with this distribution: AMRO
  E1 <-  MyData$Abundance-MyData$Fitted
}

# Check for overdispersion
N <- nrow(MyData)
p <- nrow(I1$summary.fixed)
Dispersion <- sum(E1^2) / (N - p)
Dispersion
I1$dic$dic

# Day of Year Trends
DayY <- I1$summary.random$day_of_survey_year
ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = DayY, aes(x = ID, y = mean)) +
  geom_ribbon(data = DayY, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15)) +
  theme(axis.title = element_text(size = 20))

```

```{r survey intervals}

intervals <- NA * (nrow(survey) - 1)
for (i in 2:nrow(survey)){
  intervals[i - 1] = survey$observation_date[i] - survey$observation_date[i - 1]
}

intervals_df <- data.frame(stat = names(summary(intervals)),
                           value = as.numeric(summary(intervals)))
x <- datatable(intervals_df)
x

ggsave()

```

```{r bird presence during surveys}

species_presence <- data %>% 
  filter(category == "species") %>% 
  group_by(species) %>% 
  summarise(num_surveys = n_distinct(observation_date)) %>% 
  arrange(desc(num_surveys))

datatable(species_presence)

over_half <- species_presence %>% 
  filter(num_surveys >= 49)

```

```{r analysis of precipitation and waterfowl}

monthly_precip <- precip %>% 
  group_by(survey_id) %>% 
  summarise(precipitation = sum(p01m,
                                na.rm = T))

add_survey_id <- data %>% 
  mutate(survey_id = year(observation_date) * 12 + month(observation_date)) %>% 
  mutate(survey_id = as.integer(survey_id - min(survey_id) + 1))

waterfowl_metrics1 <- add_survey_id %>% 
  filter(general_type == "Waterfowl & Friends") %>% 
  group_by(survey_id) %>% 
  summarise(abundance = sum(count),
            richness = length(unique(species)))

# adding survey ids 48 and 60, which did not record any "waterfowl and friends"
waterfowl_metrics2 <- rbind(waterfowl_metrics1,data.frame(survey_id = c(48,60),
                                                          abundance = c(0,0),
                                                          richness = c(0,0)))

### abundance

ggplot() + 
  geom_line(data = waterfowl_metrics2,
            aes(x = survey_id,
                y = abundance,
                color = "Abundance")) + 
  geom_line(data = monthly_precip,
            aes(x = survey_id,
                y = precipitation,
                color = "Precipitation")) + 
  scale_color_manual(
    values = c("Abundance" = "sienna4",
               "Precipitation" = "mediumblue")) + 
  theme_light() + 
  labs(x = "Survey ID",
       color = NULL) + 
  scale_x_continuous(breaks = seq(0,96,
                                  by = 12)) + 
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,600),
                     name = "Total Abundance",
                     sec.axis = sec_axis(~ .,
                                         name = "Total Precipitation (mm)")) + 
  theme(legend.position = "bottom")

### richness

scaling_factor <- max(waterfowl_metrics2$richness,na.rm = T) / 
  max(monthly_precip$precipitation,na.rm = T)

ggplot() + 
  geom_line(data = waterfowl_metrics2,
            aes(x = survey_id,
                y = richness,
                color = "Richness")) + 
  geom_line(data = monthly_precip,
            aes(x = survey_id,
                y = precipitation * scaling_factor,
                color = "Precipitation")) + 
  scale_color_manual(
    values = c("Richness" = "sienna4",
               "Precipitation" = "mediumblue")) + 
  theme_light() + 
  labs(x = "Survey ID",
       color = NULL) + 
  scale_x_continuous(breaks = seq(0,96,
                                  by = 12)) + 
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,20),
                     name = "Species Richness",
                     sec.axis = sec_axis(~ . / scaling_factor,
                                         name = "Total Precipitation (mm)")) + 
  theme(legend.position = "bottom")

```


Discussion



Archived visualizations (ignore)

```{r}
#| label: fig-bar
#| message: false
#| warning: false
#| fig-cap: Species richness per survey year at NCOS. The lowest richness occurs at survey year 1 with the rest appearing more similar.

# notably, no "spuh" category observations

richness <- data %>% 
  filter(!survey_year == 9, category == "species") %>% 
  group_by(survey_year) %>% 
  summarise(n = length(unique(species)))

ggplot(richness,
       aes(x = factor(survey_year),
           y = n,
           fill = survey_year)) + 
  geom_col() +
  geom_text(aes(x = survey_year, y = n + 5, label = n)) + 
  theme_classic() + 
  scale_fill_viridis_c(option = "C",
                       alpha = 0.85,
                       end = 0.9) + 
  labs(title = "Species Richness by Survey Year",
       x = "Survey Year",
       y = "Number of Species Detected") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 14),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        legend.position = "none") + 
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,200, by = 25),
                     limits = c(0,150))

```

```{r}

ebird_sy <- ncos_ebird %>% # sy = survey year
  mutate(survey_year = case_when(
      observation_date >= as.Date("2017-09-01") & observation_date <= as.Date("2018-08-31") ~ 1,
      observation_date >= as.Date("2018-09-01") & observation_date <= as.Date("2019-08-31") ~ 2,
      observation_date >= as.Date("2019-09-01") & observation_date <= as.Date("2020-08-31") ~ 3,
      observation_date >= as.Date("2020-09-01") & observation_date <= as.Date("2021-08-31") ~ 4,
      observation_date >= as.Date("2021-09-01") & observation_date <= as.Date("2022-08-31") ~ 5,
      observation_date >= as.Date("2022-09-01") & observation_date <= as.Date("2023-08-31") ~ 6,
      observation_date >= as.Date("2023-09-01") & observation_date <= as.Date("2024-08-31") ~ 7,
      observation_date >= as.Date("2024-09-01") & observation_date <= as.Date("2025-08-31") ~ 8,
      .default = NA))

ebird_richness <- ebird_sy %>% 
  filter(category == "species",
         !is.na(survey_year)) %>% 
  group_by(survey_year) %>% 
  summarize(richness = length(unique(common_name)))

ggplot(ebird_richness,
       aes(x = survey_year,
           y = richness,
           fill = survey_year)) + 
  geom_col() + 
  geom_text(aes(x = survey_year,
                y = richness + 7,
                label = richness)) + 
  theme_classic() + 
  scale_fill_viridis_c(option = "C",
                  alpha = 0.85,
                  end = 0.9) + 
  labs(title = "eBird NCOS Detected Species Richness vs Survey Year",
       x = "Survey Year",
       y = "Species Richness") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 14,
                                  face = "bold"),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        legend.position = "none") + 
  scale_x_continuous(breaks = seq(1,8,by = 1)) + 
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,225,by = 25),
                     limits = c(0,225))

```

```{r}

richness_composite <- inner_join(richness,ebird_richness) %>% 
  mutate(diff = richness - n)

ggplot(richness_composite,
       aes(x = survey_year,
           y = diff,
           fill = survey_year)) + 
  geom_col() + 
  geom_text(aes(x = survey_year,
                y = diff + 5,
                label = diff)) + 
  theme_classic() + 
  scale_fill_viridis_c(option = "C",
                       alpha = 0.85,
                       end = 0.9) + 
  labs(title = "Difference in NCOS Species Richness Between eBird and CCBER",
       x = "Survey Year",
       y = "Species Richness Difference (eBird - CCBER)") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 10),
        axis.title.x = element_text(size = 13),
        axis.title.y = element_text(size = 9),
        legend.position = "none") + 
  scale_x_continuous(breaks = seq(1,8,by = 1)) + 
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,100,by = 20),
                     limit = c(0,100))

```

```{r visualizing U values from wright-ueda}

u_path <- here("data","Matt_Vinh","wright-ueda_u_values.csv")
u_data <- read_csv(u_path,
                   col_names = F)

ggplot(u_data,
       aes(x = X1)) + 
  geom_histogram()

summary(u_data$X1)

```

```{r maximum survey count per survey year nmds}
#| message: false

# data formatting

year_max <- data %>% 
  group_by(species,survey_year) %>% 
  summarise(max_count = max(count))
top_species_pyear <- year_max %>% 
  group_by(survey_year) %>% 
  slice_max(order_by = max_count,n = 3)

community2 <- xtabs(max_count ~ survey_year + species, data = year_max)
community2 <- as.matrix(unclass(community2))

# nmds calculation

nmds2 <- metaMDS(community2,
                 autotransform = F,
                 distance = "bray",
                 engine = "monoMDS",
                 k = 2,
                 weakties = T,
                 model = "global",
                 maxit = 300,
                 try = 40,
                 trymax = 100,
                 trace = F)

nmds2_site_scores <- as.data.frame(scores(nmds2)$sites) %>% 
  rownames_to_column(var = "survey_year")
nmds2_site_scores$survey_year <- as.numeric(nmds2_site_scores$survey_year)
nmds2_site_scores <- nmds2_site_scores %>% 
  arrange(survey_year) %>% 
  mutate(xend = lead(NMDS1),yend = lead(NMDS2))

nmds2_species_scores <- as.data.frame(scores(nmds2,"species"))
nmds2_species_scores$species <- rownames(nmds2_species_scores)

nmds2_species_scores_joined <- right_join(nmds2_species_scores, top_species_pyear) %>% 
  distinct(species, .keep_all = T)
nmds2_species_scores_joined$survey_year <- as.numeric(nmds2_species_scores_joined$survey_year)

# visualization

nmdsfigure2 <- ggplot() + 
  geom_point(data = nmds2_site_scores,
                      aes(x = NMDS1, y = NMDS2,
                          color = survey_year),
             size = 5) + 
  geom_segment(data = nmds2_site_scores,
               aes(x = NMDS1, y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.35,"cm")),
               linewidth = 1.25) + 
  geom_text_repel(data = nmds2_species_scores_joined,
                  aes(x = NMDS1, y = NMDS2,
                      label = species),
                  max.overlaps = 500,size = 5.5) + 
  annotate("text",
           x = -0.275, y = 0.1,
           label = paste("2D stress =", round(nmds2$stress,3)),
           size = 6) + 
  scale_color_viridis_c(option = "C",alpha = 0.8) + 
  theme_light() + 
  labs(color = "Survey Year") + 
  theme(axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 15),
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 15),
        legend.position = c(0.15, 0.15),
        text = element_text(family = "Avenir LT Std"))

# saving

ggsave(paste0("Maximum_Species_Count_per_Survey_Year_NMDS_",
              format(Sys.Date(), "%y-%m-%d"),
              ".svg"),
       nmdsfigure2,
       dpi = 1000,
       width = 9,
       height = 8,
       units = "in")

```

```{r maximum survey count per survey year nmds}
#| message: false
#| warning: false

# data formatting

year_mean <- data %>% 
  group_by(species,survey_year) %>% 
  summarise(mean_count = mean(count))
top_species_pyear <- year_mean %>% 
  group_by(survey_year) %>% 
  slice_max(order_by = mean_count,n = 3)

community3 <- xtabs(mean_count ~ survey_year + species, data = year_mean)
community3 <- as.matrix(unclass(community3))

# nmds calculation

nmds3 <- metaMDS(community3,
                 autotransform = F,
                 distance = "bray",
                 engine = "monoMDS",
                 k = 2,
                 weakties = T,
                 model = "global",
                 maxit = 300,
                 try = 40,
                 trymax = 100,
                 trace = F)

nmds3_site_scores <- as.data.frame(scores(nmds3)$sites) %>% 
  rownames_to_column(var = "survey_year")
nmds3_site_scores$survey_year <- as.numeric(nmds3_site_scores$survey_year)
nmds3_site_scores <- nmds3_site_scores %>% 
  arrange(survey_year) %>% 
  mutate(xend = lead(NMDS1),yend = lead(NMDS2))

nmds3_species_scores <- as.data.frame(scores(nmds3,"species"))
nmds3_species_scores$species <- rownames(nmds3_species_scores)

nmds3_species_scores_joined <- right_join(nmds3_species_scores, top_species_pyear) %>% 
  distinct(species, .keep_all = T)
nmds3_species_scores_joined$survey_year <- as.numeric(nmds3_species_scores_joined$survey_year)

# visualization

nmdsfigure3 <- ggplot() + 
  geom_point(data = nmds3_site_scores,
                      aes(x = NMDS1, y = NMDS2,
                          color = survey_year),
             size = 3) + 
  geom_segment(data = nmds3_site_scores,
               aes(x = NMDS1, y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  geom_text_repel(data = nmds3_species_scores_joined,
                  aes(x = NMDS1, y = NMDS2,
                      label = species),
                  max.overlaps = 500,size = 3) + 
  annotate("text",
           x = -0.2, y = 0.1,
           label = paste("2D stress =", round(nmds3$stress,3))) + 
  scale_color_viridis_c(option = "C",alpha = 0.8) + 
  theme_light() + 
  labs(title = "Mean Species Abundance per Year NMDS",
       color = "Survey Year") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 15)) + 
  xlim(c(-0.3, 0.3)) + 
  ylim(c(-0.2, 0.2))

# saving

ggsave(paste0("Mean_Species_Abundance_per_Survey_Year_NMDS_",
              format(Sys.Date(), "%y-%m-%d"),
              ".pdf"),
       nmdsfigure3,
       width = 9,
       height = 8,
       units = "in")

```

```{r understanding aat species groups}

# negative trend

temp <- data %>% filter(species %in% c("Mourning Dove",
                                       "Western Meadowlark",
                                       "American Pipit",
                                       "Western Sandpiper"))
ggplot(temp,
       aes(x = substrate)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "negative abundance trend species")
# most popular substrates: bare ground, herbaceous vegetation, mud

# positive trend

temp <- data %>% filter(species %in% c("Snowy Egret",
                                       "Lesser Goldfinch",
                                       "White-crowned Sparrow",
                                       "Mallard",
                                       "Black-necked Stilt",
                                       "American Coot",
                                       "Song Sparrow",
                                       "Red-winged Blackbird",
                                       "Blue-gray Gnatcatcher",
                                       "Turkey Vulture",
                                       "Bushtit",
                                       "Common Yellowthroat",
                                       "Black-crowned Night Heron",
                                       "Savannah Sparrow"))
ggplot(temp,
       aes(x = substrate)) + 
  geom_bar() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "positive abundance trend species")
# most popular substrates: shrubs, herbaceous vegetation, trees, water (floating)

```

