---
title: "North Campus Open Space bird community analysis: survey years 1-8 (2017-2025) "
format: pdf
execute: 
  echo: false
fig-align: center
author: Matt Vinh
---

```{r setup,include = F}

library(here)
library(tidyverse)
library(vegan)
library(cowplot)

datapath <- here("data","aggregated","Matt_Vinh","compiled_and_cleaned_2026-02-10.rds")
data <- readRDS(datapath)

```

## Abstract (Executive summary)

## Introduction

## Methods

### Data collection

\[Fill this in based on previous documentation in NCOS annual reports and in Dryad data publication\]

### Data cleaning

-   summarize what happens in cleaning script

-   standardize species names according to XYZ (specific ebird taxonomy version)

### Data analysis

Species richness

Species composition (NMDS & guild trends)

Temporal trends for specific groups

-waterfowl, breeding birds (April-July, evidence of breeding in SB BBS), raptors...

Basic comparison to ebird data over same period

## Results

### Data set overview

## Species richness over time

## Comparison to ebird data

```{r}
#| label: ebird comparison
#| message: false
#| warning: false


# This rds is not synced to github... you have to download it from Google Drive
#It inncludes all ebird checklists for the NCOS hotspot... more filtering required
datapath_ebird <- here("data","ebird","ncos_data.rds")
ncos_ebird <- read_rds(datapath_ebird)

species_ebird <- ncos_ebird %>% 
  filter(category == "species") %>% 
  #FIXME- Matt, refine date range
  filter(observation_date > "2017-01-01") %>% 
  pull(common_name) %>% 
  unique()

species_monthly_surveys <- data %>% 
  filter(category == "species") %>% 
  pull(species) %>% 
  unique()

#what species are in ebird that are NOT in monthly surveys
setdiff(species_ebird, species_monthly_surveys)

#what species are in our monthly surveys but not in ebird (probably nothing!)
setdiff(species_monthly_surveys, species_ebird)

#one species! common gull. Maybe mis-identified?
```


## Species composition over time

The lowest richness occurs at survey year 1 with the following years appearing more similar.

```{r}
#| label: fig-bar
#| message: false
#| warning: false
#| fig-cap: Species richness per survey year at NCOS. The lowest richness occurs at survey year 1 with the rest appearing more similar.

# notably, no "spuh" category observations

richness <- data %>% 
  filter(!survey_year == 9, category == "species") %>% 
  group_by(survey_year) %>% 
  summarise(n = length(unique(species)))

ggplot(richness,
       aes(x = factor(survey_year),
           y = n,
           fill = survey_year)) + 
  geom_col() +
  #geom_point() +
  theme_light() + 
  theme_cowplot() +
  scale_fill_viridis_c(option = "C",
                       alpha = 0.85,
                       end = 0.9) + 
  labs(title = "Species Richness by Survey Year",
       x = "Survey Year",
       y = "Number of Species Detected") + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none") + 
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,200, by = 25),
                     limits = c(0,150)
                     ) +
  geom_text(aes(x = survey_year, y = n + 5, label = n))
  

```

```{r}
#| label: fig-nmds
#| message: false
#| warning: false
#| fig-cap: Revised NMDS based on 'Maximum Species Abundance' of each species in each year. I reproduced it thinking that with cleaner data it might appear differently, giving insight into species identity similarities or differences, following the bar chart of species richness. The usage of setdiff() may be more desirable in this situation.

year_max <- data %>% 
  filter(!survey_year == 9) %>%
  group_by(survey_year, species) %>%
  summarise(max_count = max(count))

community <- as.matrix(unclass(xtabs(max_count ~ survey_year + species,
                                     data = year_max)))

nmds <- metaMDS(community,autotransform = F,distance = "bray",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

nmds_site_scores <- as.data.frame(scores(nmds)$sites) %>% 
  rownames_to_column(var = "survey_year") %>% 
  mutate(survey_year = as.numeric(survey_year)) %>% 
  arrange(survey_year) %>% 
  mutate(
    xend = lead(NMDS1),
    yend = lead(NMDS2))

ggplot() + 
  geom_point(data = nmds_site_scores,
                      aes(x = NMDS1,
                          y = NMDS2,
                          color = survey_year),
             size = 3) + 
  geom_segment(data = nmds_site_scores,
               aes(x = NMDS1,
                   y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  annotate("text",
           x = -0.2,
           y = 0.15,
           label = paste("2D stress =",
                         round(nmds$stress,3))) + 
  scale_color_viridis_c(option = "C",
                        alpha = 0.8,
                        end = 0.95) + 
  theme_light() + 
  labs(title = "Maximum Species Abundance by Survey Year NMDS",
       color = "Survey Year") + 
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
# TODO- this chunk was a rough attempt at creating an NMDS based on presence-absence (really presence vs. non-detection) of species rather than abundance. 

nmds_binary <- metaMDS(community,autotransform = T, distance = "jaccard",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

nmds_site_scores_binary <- as.data.frame(scores(nmds_binary)$sites) %>% 
  rownames_to_column(var = "survey_year") %>% 
  mutate(survey_year = as.numeric(survey_year)) %>% 
  arrange(survey_year) %>% 
  mutate(
    xend = lead(NMDS1),
    yend = lead(NMDS2))

ggplot() + 
  geom_point(data = nmds_site_scores_binary,
                      aes(x = NMDS1,
                          y = NMDS2,
                          color = survey_year),
             size = 3) + 
  geom_segment(data = nmds_site_scores_binary,
               aes(x = NMDS1,
                   y = NMDS2,
                   xend = xend,
                   yend = yend,
                   color = survey_year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  annotate("text",
           x = -0.2,
           y = 0.15,
           label = paste("2D stress =",
                         round(nmds_binary$stress,3))) + 
  scale_color_viridis_c(option = "C",
                        alpha = 0.8,
                        end = 0.95) + 
  theme_light() + 
  labs(title = "NMDS based on Jaccard distance (presence/absence)",
       color = "Survey Year") + 
  theme(plot.title = element_text(hjust = 0.5))
```

Discussion
