---
title: "Wright-Ueda_scratchwork"
format: html
---

```{r setup}

library(here)
library(tidyverse)
library(GGally)
library(plotly)
library(vegan)
library(viridis) # color scheme
library(zoo) # for rolling average
library(roll) # for rolling sd
library(car)
library(lubridate)
library(devtools) 
library(geoR)
library(INLA)

jrbp_path <- here("data","JRBP_Bird_Transect_Data.csv")
df <- read_csv(jrbp_path)

```



In an attempt to understand the criteria used to determine hyperparameter U value selection, here all statistical analyses from the paper are replicated using three example species from the JRBP_Bird_Transet_Data.csv file.

After running everything, plots c - p return errors or NULL and it is not clear why. It is worth noting that all plots using "E1" had to be modified to instead use "E1$Abundance" due to raised errors about incorrect data input, however all plots modified this way are a portion of those that return NULL.

Everything is executed in the following chunk with the results saved to variables. Below this chunk there are individual chunks for each different output, for comparison of results between the three exemplified JRBP species.



```{r examining three species with results shown in paper}

# CASJ ####
species <- 'CASJ'
U <- 0.007
distribution <- 'nbinomial'

speciescol <- paste0('n.',species)
MyData <- na.exclude(df[, c(speciescol,
                            "Census_Area", "Day_of_year", "Date", 'Year',
                            'ID',
                            "Observer1")])
names(MyData)[1] <- 'Abundance'
bspecies <- MyData[,1]

hyper.prec <- list(theta = list(prior = "pc.prec", param = c(U, 0.05)))
inla.setOption(scale.model.default = TRUE)

I1 <- inla(Abundance ~ f(Year, model = "rw1", hyper = hyper.prec) + 
             f(Day_of_year, model = "rw2", hyper = hyper.prec) +
             f(Census_Area, model = "iid") +
             f(Observer1, model = "iid"),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)

#### Model Validation ####
MyData$Fitted <- I1$summary.fitted.values[,'mean']
# Plot trend for year
Yearsm <- I1$summary.random$Year
plot1a <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = Yearsm, aes(x = ID, y = mean)) +
  geom_ribbon(data = Yearsm, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  ggtitle('Trend by Year') +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(Yearsm$`0.025quant`)-0.05,max(Yearsm$`0.975quant`)+0.05)


### Dispersion ###
mu <- I1$summary.fitted.values[, "mean"]
theta <- I1$summary.hyperpar[1, "mean"]

if(distribution == 'nbinomial'){
  # Pearson residuals for negative binomial distribution
  E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
}
if(distribution == 'poisson'){
  # Pearson residuals for poisson distribution
  E1 <- (bspecies - mu) / sqrt(mu) 
}
if(distribution == 'gaussian'){
  # Residuals for gaussian distribution
  E1 <-  MyData$Abundance-MyData$Fitted
}

# Check for overdispersion
N <- nrow(MyData)
p <- nrow(I1$summary.fixed)
Dispersion1 <- sum(E1^2) / (N - p)
Dispersion1
value1a <- I1$dic$dic

# Day of Year Trends
DayY <- I1$summary.random$Day_of_year
plot1b <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = DayY, aes(x = ID, y = mean)) +
  geom_ribbon(data = DayY, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15)) +
  theme(axis.title = element_text(size = 20))

MyData$Residuals <- E1
plot1c <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Residuals), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 

# Variogram for checking for temporal autocorrelation by year 
breaks <- seq(0.5, 20.5, length = 20)
MyData$E1 <- E1
MyData$Zeros <- rep(0, nrow(MyData))
V1 <- variog(coords = MyData[, c("Year", "Zeros")], data = MyData$E1, breaks = breaks)
plot1d <- plot(V1, type = "b", 
     xlab = "Time difference (years)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5)

# Variogram for checking for temporal autocorrelation by data of year
breaks2 <- seq(0.5, 364.5, length = 365)
V1 <- variog(coords = MyData[, c("Day_of_year", "Zeros")], data = MyData$E1, breaks = breaks2)
plot1e <- plot(V1, type = "b", 
     xlab = "Time difference (day of year)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5, ylim = c(0,2))

# Plot Residuals vs. fitted values and covariates
plot1f <- plot(E1$Abundance ~ MyData$Fitted) # Residuals vs fitted values
plot1g <- qqnorm(E1$Abundance)
plot1h <- hist(E1$Abundance) # histogram of residuals
plot1i <- plot(E1$Abundance ~ MyData$Abundance, data = MyData) 
plot1j <- plot(MyData$Fitted ~ MyData$Abundance, asp= 1, data = MyData)
plot1k <- plot(MyData$Fitted ~ MyData$Abundance, data = MyData)

plot1l <- plot(E1$Abundance ~ Year, data = MyData)
plot1m <- plot(E1$Abundance ~ Day_of_year, data = MyData)

plot1n <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot1o <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') 
plot1p <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Species Abundance') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted abundance per transect')

plot1q <- ggplot(data = MyData) +
  geom_point(aes(x = Date,y = Fitted), colour = 'blue', alpha = 1/10) +
  geom_point(aes(x = Date,y = Abundance), colour = 'red', alpha = 1/10) +
  ylim(c(0,20))

plot1r <- ggplot(data = MyData) +
  geom_smooth(aes(x = Day_of_year ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot1s <- ggplot(data = MyData) +
  geom_point(aes(x = Day_of_year ,y = Abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = Day_of_year ,y = Fitted), 
              span = 0.2, colour = 'blue',  
              method = 'loess')


# Plot random effects for survey point
point.r <- I1$summary.random$Census_Area
plot1t <- ggplot(data = point.r) +
  geom_pointrange(aes(x = ID, y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Point") +
  ylab("Smoother")

# Plot random effects for observer 
obs.r <- I1$summary.random$Observer1
obs.r$ID <- as.factor(obs.r$ID)
obs.counts.surveys <- table(df$Observer1)
obs.order <- as.data.frame(obs.counts.surveys)
colnames(obs.order) <- c("ID",'count')

obs.r <- obs.order %>%
  left_join(obs.r, by = "ID")

# Make the plot
plot1u <- ggplot(data = obs.r ) +
  geom_pointrange(aes(x = reorder(ID, -count), y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Observer") +
  ylab("Smoother") +
  coord_flip()  +
  theme(axis.text.y = element_text(size = 5)) +
  theme(axis.text.x = element_text(size = 7))

#### Percent Change in Abundance ####

# Extract the posterior marginal distributions for Year random walk
year.marg <- I1$marginals.random$Year 
value1b <- year.marg[[32]]

# Create dataframe for storing data
Magdf <- data.frame(Year = seq(from = 1989, to = 2020, by = 1))
for(i in 1:32) { 
  # Transform the posterior marginal distribution for current year so that it is not in log scale
  tmarg <- inla.tmarginal(exp, year.marg[[i]]) 
  # Calculate mean value for year i
  Magdf$mean[i] <- mean(tmarg[,1]) 
  # Calculate 95% CRI for year i's estimate
  Magdf$cri95[i] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                         inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
  # Calculate mean percent change current year vs initial
  Magdf$percent.change.mean[i] <- 100 * (Magdf$mean[i] - Magdf$mean[1]) / Magdf$mean[1] 
  # Calculate variance in percent change using 95% CRIs and variance estimation method 
  Magdf$percent.change.cri95[i] <- Magdf$mean[i] / Magdf$mean[1] *
    sqrt((Magdf$cri95[i]^2 / Magdf$mean[i]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
  Magdf$lower.cri95[i] <- Magdf$percent.change.mean[i] - (Magdf$percent.change.cri95[i])
  Magdf$upper.cri95[i] <- Magdf$percent.change.mean[i] + (Magdf$percent.change.cri95[i])
}

# Mean percent change since start of survey
value1c <- Magdf$percent.change.mean[32] 

# 95% CRI for percent change since start of survey
value1d <- Magdf$percent.change.cri95[32]

# Both of the above values were recorded in the Species Organization Dryad File
# for every species.

# Plot percent change since 1999 and 95% CRIs for all years
plot1v <- ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance since 1989") +
  geom_line(data = Magdf, aes(x = Year, y = percent.change.mean), color = "black") +
  geom_ribbon(data = Magdf, aes(x = Year, ymin = lower.cri95, ymax = upper.cri95), alpha = 0.3) +
  xlim(1989,2020) # constrain the x axis so that the starting year is not showing

# CBCH ####
species <- 'CBCH'
U <- 0.012 
distribution <- 'nbinomial'

speciescol <- paste0('n.',species)
MyData <- na.exclude(df[, c(speciescol,
                            "Census_Area", "Day_of_year", "Date", 'Year',
                            'ID',
                            "Observer1")])
names(MyData)[1] <- 'Abundance'
bspecies <- MyData[,1]

hyper.prec <- list(theta = list(prior = "pc.prec", param = c(U, 0.05)))
inla.setOption(scale.model.default = TRUE)

I1 <- inla(Abundance ~ f(Year, model = "rw1", hyper = hyper.prec) + 
             f(Day_of_year, model = "rw2", hyper = hyper.prec) +
             f(Census_Area, model = "iid") +
             f(Observer1, model = "iid"),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)

#### Model Validation ####
MyData$Fitted <- I1$summary.fitted.values[,'mean']
# Plot trend for year
Yearsm <- I1$summary.random$Year
plot2a <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = Yearsm, aes(x = ID, y = mean)) +
  geom_ribbon(data = Yearsm, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  ggtitle('Trend by Year') +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(Yearsm$`0.025quant`)-0.05,max(Yearsm$`0.975quant`)+0.05)


### Dispersion ###
mu <- I1$summary.fitted.values[, "mean"]
theta <- I1$summary.hyperpar[1, "mean"]

if(distribution == 'nbinomial'){
  # Pearson residuals for negative binomial distribution
  E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
}
if(distribution == 'poisson'){
  # Pearson residuals for poisson distribution
  E1 <- (bspecies - mu) / sqrt(mu) 
}
if(distribution == 'gaussian'){
  # Residuals for gaussian distribution
  E1 <-  MyData$Abundance-MyData$Fitted
}

# Check for overdispersion
N <- nrow(MyData)
p <- nrow(I1$summary.fixed)
Dispersion2 <- sum(E1^2) / (N - p)
Dispersion2
value2a <- I1$dic$dic

# Day of Year Trends
DayY <- I1$summary.random$Day_of_year
plot2b <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = DayY, aes(x = ID, y = mean)) +
  geom_ribbon(data = DayY, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15)) +
  theme(axis.title = element_text(size = 20))

MyData$Residuals <- E1
plot2c <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Residuals), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 

# Variogram for checking for temporal autocorrelation by year 
breaks <- seq(0.5, 20.5, length = 20)
MyData$E1 <- E1
MyData$Zeros <- rep(0, nrow(MyData))
V1 <- variog(coords = MyData[, c("Year", "Zeros")], data = MyData$E1, breaks = breaks)
plot2d <- plot(V1, type = "b", 
     xlab = "Time difference (years)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5)

# Variogram for checking for temporal autocorrelation by data of year
breaks2 <- seq(0.5, 364.5, length = 365)
V1 <- variog(coords = MyData[, c("Day_of_year", "Zeros")], data = MyData$E1, breaks = breaks2)
plot2e <- plot(V1, type = "b", 
     xlab = "Time difference (day of year)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5, ylim = c(0,2))

# Plot Residuals vs. fitted values and covariates
plot2f <- plot(E1$Abundance ~ MyData$Fitted) # Residuals vs fitted values
plot2g <- qqnorm(E1$Abundance)
plot2h <- hist(E1$Abundance) # histogram of residuals
plot2i <- plot(E1$Abundance ~ MyData$Abundance, data = MyData) 
plot2j <- plot(MyData$Fitted ~ MyData$Abundance, asp= 1, data = MyData)
plot2k <- plot(MyData$Fitted ~ MyData$Abundance, data = MyData)

plot2l <- plot(E1$Abundance ~ Year, data = MyData)
plot2m <- plot(E1$Abundance ~ Day_of_year, data = MyData)

plot2n <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot2o <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') 
plot2p <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Species Abundance') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted abundance per transect')

plot2q <- ggplot(data = MyData) +
  geom_point(aes(x = Date,y = Fitted), colour = 'blue', alpha = 1/10) +
  geom_point(aes(x = Date,y = Abundance), colour = 'red', alpha = 1/10) +
  ylim(c(0,20))

plot2r <- ggplot(data = MyData) +
  geom_smooth(aes(x = Day_of_year ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot2s <- ggplot(data = MyData) +
  geom_point(aes(x = Day_of_year ,y = Abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = Day_of_year ,y = Fitted), 
              span = 0.2, colour = 'blue',  
              method = 'loess')


# Plot random effects for survey point
point.r <- I1$summary.random$Census_Area
plot2t <- ggplot(data = point.r) +
  geom_pointrange(aes(x = ID, y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Point") +
  ylab("Smoother")

# Plot random effects for observer 
obs.r <- I1$summary.random$Observer1
obs.r$ID <- as.factor(obs.r$ID)
obs.counts.surveys <- table(df$Observer1)
obs.order <- as.data.frame(obs.counts.surveys)
colnames(obs.order) <- c("ID",'count')

obs.r <- obs.order %>%
  left_join(obs.r, by = "ID")

# Make the plot
plot2u <- ggplot(data = obs.r ) +
  geom_pointrange(aes(x = reorder(ID, -count), y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Observer") +
  ylab("Smoother") +
  coord_flip()  +
  theme(axis.text.y = element_text(size = 5)) +
  theme(axis.text.x = element_text(size = 7))

#### Percent Change in Abundance ####

# Extract the posterior marginal distributions for Year random walk
year.marg <- I1$marginals.random$Year 
value2b <- year.marg[[32]]

# Create dataframe for storing data
Magdf <- data.frame(Year = seq(from = 1989, to = 2020, by = 1))
for(i in 1:32) { 
  # Transform the posterior marginal distribution for current year so that it is not in log scale
  tmarg <- inla.tmarginal(exp, year.marg[[i]]) 
  # Calculate mean value for year i
  Magdf$mean[i] <- mean(tmarg[,1]) 
  # Calculate 95% CRI for year i's estimate
  Magdf$cri95[i] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                         inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
  # Calculate mean percent change current year vs initial
  Magdf$percent.change.mean[i] <- 100 * (Magdf$mean[i] - Magdf$mean[1]) / Magdf$mean[1] 
  # Calculate variance in percent change using 95% CRIs and variance estimation method 
  Magdf$percent.change.cri95[i] <- Magdf$mean[i] / Magdf$mean[1] *
    sqrt((Magdf$cri95[i]^2 / Magdf$mean[i]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
  Magdf$lower.cri95[i] <- Magdf$percent.change.mean[i] - (Magdf$percent.change.cri95[i])
  Magdf$upper.cri95[i] <- Magdf$percent.change.mean[i] + (Magdf$percent.change.cri95[i])
}

# Mean percent change since start of survey
value2c <- Magdf$percent.change.mean[32] 

# 95% CRI for percent change since start of survey
value2d <- Magdf$percent.change.cri95[32]

# Both of the above values were recorded in the Species Organization Dryad File
# for every species.

# Plot percent change since 1999 and 95% CRIs for all years
plot2v <- ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance since 1989") +
  geom_line(data = Magdf, aes(x = Year, y = percent.change.mean), color = "black") +
  geom_ribbon(data = Magdf, aes(x = Year, ymin = lower.cri95, ymax = upper.cri95), alpha = 0.3) +
  xlim(1989,2020) # constrain the x axis so that the starting year is not showing

# RTHA ####
species <- 'RTHA'
U <- 0.02
distribution <- 'poisson'

speciescol <- paste0('n.',species)
MyData <- na.exclude(df[, c(speciescol,
                            "Census_Area", "Day_of_year", "Date", 'Year',
                            'ID',
                            "Observer1")])
names(MyData)[1] <- 'Abundance'
bspecies <- MyData[,1]

hyper.prec <- list(theta = list(prior = "pc.prec", param = c(U, 0.05)))
inla.setOption(scale.model.default = TRUE)

I1 <- inla(Abundance ~ f(Year, model = "rw1", hyper = hyper.prec) + 
             f(Day_of_year, model = "rw2", hyper = hyper.prec) +
             f(Census_Area, model = "iid") +
             f(Observer1, model = "iid"),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)

#### Model Validation ####
MyData$Fitted <- I1$summary.fitted.values[,'mean']
# Plot trend for year
Yearsm <- I1$summary.random$Year
plot3a <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = Yearsm, aes(x = ID, y = mean)) +
  geom_ribbon(data = Yearsm, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  ggtitle('Trend by Year') +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(Yearsm$`0.025quant`)-0.05,max(Yearsm$`0.975quant`)+0.05)


### Dispersion ###
mu <- I1$summary.fitted.values[, "mean"]
theta <- I1$summary.hyperpar[1, "mean"]

if(distribution == 'nbinomial'){
  # Pearson residuals for negative binomial distribution
  E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
}
if(distribution == 'poisson'){
  # Pearson residuals for poisson distribution
  E1 <- (bspecies - mu) / sqrt(mu) 
}
if(distribution == 'gaussian'){
  # Residuals for gaussian distribution
  E1 <-  MyData$Abundance-MyData$Fitted
}

# Check for overdispersion
N <- nrow(MyData)
p <- nrow(I1$summary.fixed)
Dispersion3 <- sum(E1^2) / (N - p)
Dispersion3
value3a <- I1$dic$dic

# Day of Year Trends
DayY <- I1$summary.random$Day_of_year
plot3b <- ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = DayY, aes(x = ID, y = mean)) +
  geom_ribbon(data = DayY, aes(x = ID, ymax = `0.025quant`, ymin = `0.975quant`), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 15)) +
  theme(axis.text.x = element_text(size = 15)) +
  theme(axis.title = element_text(size = 20))

MyData$Residuals <- E1
plot3c <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Residuals), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 

# Variogram for checking for temporal autocorrelation by year 
breaks <- seq(0.5, 20.5, length = 20)
MyData$E1 <- E1
MyData$Zeros <- rep(0, nrow(MyData))
V1 <- variog(coords = MyData[, c("Year", "Zeros")], data = MyData$E1, breaks = breaks)
plot3d <- plot(V1, type = "b", 
     xlab = "Time difference (years)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5)

# Variogram for checking for temporal autocorrelation by data of year
breaks2 <- seq(0.5, 364.5, length = 365)
V1 <- variog(coords = MyData[, c("Day_of_year", "Zeros")], data = MyData$E1, breaks = breaks2)
plot3e <- plot(V1, type = "b", 
     xlab = "Time difference (day of year)", ylab = "Sample Variogram",
     col = 1, pch = 16, cex = 1.5, ylim = c(0,2))

# Plot Residuals vs. fitted values and covariates
plot3f <- plot(E1$Abundance ~ MyData$Fitted) # Residuals vs fitted values
plot3g <- qqnorm(E1$Abundance)
plot3h <- hist(E1$Abundance) # histogram of residuals
plot3i <- plot(E1$Abundance ~ MyData$Abundance, data = MyData) 
plot3j <- plot(MyData$Fitted ~ MyData$Abundance, asp= 1, data = MyData)
plot3k <- plot(MyData$Fitted ~ MyData$Abundance, data = MyData)

plot3l <- plot(E1$Abundance ~ Year, data = MyData)
plot3m <- plot(E1$Abundance ~ Day_of_year, data = MyData)

plot3n <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot3o <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') 
plot3p <- ggplot(data = MyData) +
  geom_smooth(aes(x = Date ,y = Fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Species Abundance') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted abundance per transect')

plot3q <- ggplot(data = MyData) +
  geom_point(aes(x = Date,y = Fitted), colour = 'blue', alpha = 1/10) +
  geom_point(aes(x = Date,y = Abundance), colour = 'red', alpha = 1/10) +
  ylim(c(0,20))

plot3r <- ggplot(data = MyData) +
  geom_smooth(aes(x = Day_of_year ,y = Abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') 
plot3s <- ggplot(data = MyData) +
  geom_point(aes(x = Day_of_year ,y = Abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = Day_of_year ,y = Fitted), 
              span = 0.2, colour = 'blue',  
              method = 'loess')


# Plot random effects for survey point
point.r <- I1$summary.random$Census_Area
plot3t <- ggplot(data = point.r) +
  geom_pointrange(aes(x = ID, y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Point") +
  ylab("Smoother")

# Plot random effects for observer 
obs.r <- I1$summary.random$Observer1
obs.r$ID <- as.factor(obs.r$ID)
obs.counts.surveys <- table(df$Observer1)
obs.order <- as.data.frame(obs.counts.surveys)
colnames(obs.order) <- c("ID",'count')

obs.r <- obs.order %>%
  left_join(obs.r, by = "ID")

# Make the plot
plot3u <- ggplot(data = obs.r ) +
  geom_pointrange(aes(x = reorder(ID, -count), y = mean, ymin = `0.025quant`, ymax = `0.975quant`)) +
  geom_hline(aes(yintercept = 0)) +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Observer") +
  ylab("Smoother") +
  coord_flip()  +
  theme(axis.text.y = element_text(size = 5)) +
  theme(axis.text.x = element_text(size = 7))

#### Percent Change in Abundance ####

# Extract the posterior marginal distributions for Year random walk
year.marg <- I1$marginals.random$Year 
value3b <- year.marg[[32]]

# Create dataframe for storing data
Magdf <- data.frame(Year = seq(from = 1989, to = 2020, by = 1))
for(i in 1:32) { 
  # Transform the posterior marginal distribution for current year so that it is not in log scale
  tmarg <- inla.tmarginal(exp, year.marg[[i]]) 
  # Calculate mean value for year i
  Magdf$mean[i] <- mean(tmarg[,1]) 
  # Calculate 95% CRI for year i's estimate
  Magdf$cri95[i] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                         inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
  # Calculate mean percent change current year vs initial
  Magdf$percent.change.mean[i] <- 100 * (Magdf$mean[i] - Magdf$mean[1]) / Magdf$mean[1] 
  # Calculate variance in percent change using 95% CRIs and variance estimation method 
  Magdf$percent.change.cri95[i] <- Magdf$mean[i] / Magdf$mean[1] *
    sqrt((Magdf$cri95[i]^2 / Magdf$mean[i]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
  Magdf$lower.cri95[i] <- Magdf$percent.change.mean[i] - (Magdf$percent.change.cri95[i])
  Magdf$upper.cri95[i] <- Magdf$percent.change.mean[i] + (Magdf$percent.change.cri95[i])
}

# Mean percent change since start of survey
value3c <- Magdf$percent.change.mean[32] 

# 95% CRI for percent change since start of survey
value3d <- Magdf$percent.change.cri95[32]

# Both of the above values were recorded in the Species Organization Dryad File
# for every species.

# Plot percent change since 1999 and 95% CRIs for all years
plot3v <- ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance since 1989") +
  geom_line(data = Magdf, aes(x = Year, y = percent.change.mean), color = "black") +
  geom_ribbon(data = Magdf, aes(x = Year, ymin = lower.cri95, ymax = upper.cri95), alpha = 0.3) +
  xlim(1989,2020) # constrain the x axis so that the starting year is not showing

```

```{r plot trend for year}
plot1a
plot2a
plot3a

# appears exactly like as in the paper on page 33
```

```{r dispersion}
Dispersion1
Dispersion2
Dispersion3

# closer to 1 = better
# under 1 = underdispersed
# over 1 = overdispersed
# since model is tuned according to dispersion via underlying distribution,
# the dispersions are close to 1?
```

```{r inla model dic}
value1a
value2a
value3a

# deviance information criterion
# used to compareto other species dic values?
```

```{r day of year trends 1}
plot1b
plot2b
plot3b
```

```{r day of year trends 2}
plot1c
plot2c
plot3c
```

```{r Variogram for checking for temporal autocorrelation by year}
plot1d
plot2d
plot3d
```

```{r Variogram for checking for temporal autocorrelation by data of year}
plot1e
plot2e
plot3e
```

```{r Residuals vs fitted values}
plot1f
plot2f
plot3f
```

```{r Plot Residuals vs. fitted values and covariates qqnorm}
plot1g
plot2g
plot3g
```

```{r Plot Residuals vs. fitted values and covariates histogram}
plot1h
plot2h
plot3h
```

```{r Plot Residuals vs. fitted values and covariates 1}
plot1i
plot2i
plot3i
```

```{r Plot Residuals vs. fitted values and covariates 2}
plot1j
plot2j
plot3j
```

```{r Plot Residuals vs. fitted values and covariates 3}
plot1k
plot2k
plot3k
```

```{r Plot Residuals vs. fitted values and covariates 4}
plot1l
plot2l
plot3l
```

```{r Plot Residuals vs. fitted values and covariates 5}
plot1m
plot2m
plot3m
```

```{r Plot Residuals vs. fitted values and covariates 6}
plot1n
plot2n
plot3n
```

```{r Plot Residuals vs. fitted values and covariates 7}
plot1o
plot2o
plot3o
```

```{r Plot Residuals vs. fitted values and covariates 8}
plot1p
plot2p
plot3p
```

```{r Plot Residuals vs. fitted values and covariates 9}
plot1q
plot2q
plot3q
```

```{r Plot Residuals vs. fitted values and covariates 10}
plot1r
plot2r
plot3r
```

```{r Plot Residuals vs. fitted values and covariates 11}
plot1s
plot2s
plot3s
```

```{r Plot random effects for survey point}
plot1t
plot2t
plot3t
```

```{r Plot random effects for observer}
plot1u
plot2u
plot3u
```

```{r Extract the posterior marginal distributions for Year random walk}
value1b
value2b
value3b
```

```{r Mean percent change since start of survey}
value1c
value2c
value3c
```

```{r 95% CRI for percent change since start of survey}
value1d
value2d
value3d
```

```{r Plot percent change since 1999 and 95% CRIs for all years}
plot1v
plot2v
plot3v
```