---
title: "Wright-Ueda Methods Replication for NCOS Bird Survey Data"
author: "Matt Vinh"
date: "2026-04-07"
format: html
---

```{r setup}
#| include: false

library(here)
library(tidyverse)
library(GGally)
library(plotly)
library(vegan)
library(viridis) # color scheme
library(zoo) # for rolling average
library(roll) # for rolling sd
library(car)
library(lubridate)
library(devtools) 
library(geoR)
library(INLA)
library(purrr)
library(knitr)

# JRBP data
jrbp_path <- here("data","JRBP_Bird_Transect_Data.csv")
df <- read_csv(jrbp_path)

# CCBER data
ccber_path <- here("data","Matt_Vinh","survey-level_2026-03-01.rds")
data <- read_rds(ccber_path)

```

Introduction

When applying the methods of Julien and Tyler to CCBER bird survey data collected at NCOS, it was quickly noticeable that one default U hyperparameter value had wildly different impact on different species. Thus, here, referencing the JRBP data and determined U values, I attempt to understand the criteria used to judge possible U values to select a specialized number for each example NCOS species examined.

## JRBP Data Testing

For my evaluation of Julien and Tyler's methods used to select U values, I examined all of their "diagnostic plots" while inputting the correct U values and population trend distributions per species simultaneously using a modified version of their provided R script. I do not include all of the visualizations they used here, instead providing a selection of the different types. It should be noted that my current implementation of modified script is not vectorized with purrr, which remains an area to be improved upon later.

```{r replicating model fitting for wright-ueda example species}
#| message: false
#| include: false

### setting desired NCOS species

species_options <- c("NOMO", # northern mockingbird, u = 0.07, nbinomial
                     "CASJ", # california scrub-jay, u = 0.007, nbinomial
                     "CAQU", # california quail, u = 0.022, nbinomial
                     "SPTO", # spotted towhee, u = 0.007, nbinomial
                     "RTHA", # red-tailed hawk, u = 0.02, poisson
                     "CBCH") # chestnut-backed chickadee, u = 0.012, nbinomial



### initialization

yearsm_list <- list()
dayY_list <- list()
myData_list <- list()
magdf_list <- list()
E1_list <- list()
dispersion_vector <- vector()
DIC_vector <- vector()
v1_list <- list()
v2_list <- list()
year_marg_list <- list()
percent_change_vector <- vector()
percent_change_intervals_vector <- vector()

breaks <- seq(0.5, 20.5, length = 20)
breaks2 <- seq(0.5, 364.5, length = 365)



for (i in 1:length(species_options)){
  
  ### INLA model
  
    species <- species_options[i]
    
    if (species == "NOMO"){U <- 0.07}
    if (species == "CASJ" | species == "SPTO"){U <- 0.007}
    if (species == "CAQU"){U <- 0.022}
    if (species == "RTHA"){U <- 0.02}
    if (species == "CBCH"){U <- 0.012}
    
    if (species == "RTHA"){distribution <- "poisson"}
    else {distribution <- "nbinomial"}

    speciescol <- paste0('n.',species)
    MyData <- na.exclude(df[, c(speciescol,
                                  "Census_Area", "Day_of_year", "Date", 'Year',
                            'ID',
                            "Observer1")])
    names(MyData)[1] <- 'Abundance'
    bspecies <- MyData[,1]

    hyper.prec <- list(theta = list(prior = "pc.prec",
                                    param = c(U, 0.05)))
    inla.setOption(scale.model.default = TRUE)

    I1 <- inla(Abundance ~ f(Year, model = "rw1", hyper = hyper.prec) + 
              f(Day_of_year, model = "rw2", hyper = hyper.prec) + 
              f(Census_Area, model = "iid") +
              f(Observer1, model = "iid"),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)
    
  ### calculations and saving data
    
    MyData$Fitted <- I1$summary.fitted.values[,'mean']
    
    # trend for year
    Yearsm <- I1$summary.random$Year
    
    # calculating residuals
    mu <- MyData$Fitted
    theta <- I1$summary.hyperpar[1, "mean"]
    
    if(distribution == 'nbinomial'){
      # Pearson residuals for negative binomial distribution
      E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
    }
    if(distribution == 'poisson'){
      # Pearson residuals for poisson distribution
      E1 <- (bspecies - mu) / sqrt(mu) 
    }
    if(distribution == 'gaussian'){
      # Residuals for gaussian distribution
      E1 <-  MyData$Abundance-MyData$Fitted
    }
    names(E1)[1] <- "Residuals"
    
    
    ### Check for overdispersion
    
    N <- nrow(MyData)
    p <- nrow(I1$summary.fixed)
    Dispersion <- sum(E1^2) / (N - p)
    
    
    ### Day of Year Trends
    
    DayY <- I1$summary.random$Day_of_year
    MyData$Residuals <- E1$Residuals
    
    
    ### variograms
    
    # Variogram for checking for temporal autocorrelation by year 
    temp <- E1$Residuals
    MyData$E1 <- temp
    MyData$Zeros <- rep(0, nrow(MyData))
    V1 <- variog(coords = MyData[, c("Year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks)
    
    # Variogram for checking for temporal autocorrelation by data of year
    V2 <- variog(coords = MyData[, c("Day_of_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks2)
    
    
    ### Extract the posterior marginal distributions for Year random walk
    year.marg <- I1$marginals.random$Year 
    
    
    ### Create dataframe for storing data
    Magdf <- data.frame(Year = seq(from = 1989, to = 2020, by = 1))
    for(j in 1:32) { 
      # Transform the posterior marginal distribution for current year so that it is not in log scale
      tmarg <- inla.tmarginal(exp, year.marg[[j]]) 
      # Calculate mean value for year i
      Magdf$mean[j] <- mean(tmarg[,1]) 
      # Calculate 95% CRI for year i's estimate
      Magdf$cri95[j] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                             inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
      # Calculate mean percent change current year vs initial
      Magdf$percent.change.mean[j] <- 100 * (Magdf$mean[j] - Magdf$mean[1]) / Magdf$mean[1] 
      # Calculate variance in percent change using 95% CRIs and variance estimation method 
      Magdf$percent.change.cri95[j] <- Magdf$mean[j] / Magdf$mean[1] *
        sqrt((Magdf$cri95[j]^2 / Magdf$mean[j]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
      Magdf$lower.cri95[j] <- Magdf$percent.change.mean[j] - (Magdf$percent.change.cri95[j])
      Magdf$upper.cri95[j] <- Magdf$percent.change.mean[j] + (Magdf$percent.change.cri95[j])
    }
    
    
    ### saving data
    
    yearsm_individual <- data.frame(species = species_options[i],
                                    ID = Yearsm$ID,
                                    mean = Yearsm$mean,
                                    quant0.025 = Yearsm$`0.025quant`,
                                    quant0.975 = Yearsm$`0.975quant`)
    
    dayY_individual <- data.frame(species = species_options[i],
                                  ID = DayY$ID,
                                  mean = DayY$mean,
                                  quant0.025 = DayY$`0.025quant`,
                                  quant0.975 = DayY$`0.975quant`)
    
    myData_individual <- data.frame(species = species_options[i],
                                    abundance = MyData$Abundance,
                                    day_of_year = MyData$Day_of_year,
                                    date = MyData$Date,
                                    year = MyData$Year,
                                    fitted = MyData$Fitted,
                                    residuals = MyData$Residuals,
                                    E1 = MyData$E1)
    
    magdf_individual <- data.frame(species = species_options[i],
                                   year = Magdf$Year,
                                   change = Magdf$percent.change.mean,
                                   lower = Magdf$lower.cri95,
                                   upper = Magdf$upper.cri95)
    
    E1_individual <- data.frame(species = species_options[i],
                                residuals = E1$Residuals)
    
    v1_individual <- data.frame(species = species_options[i],
                                u = V1$u,
                                v = V1$v)
    
    v2_individual <- data.frame(species = species_options[i],
                                u = V2$u,
                                v = V2$v)
    
  # data frames
  yearsm_list[[i]] <- yearsm_individual
  dayY_list[[i]] <- dayY_individual
  myData_list[[i]] <- myData_individual
  magdf_list[[i]] <- magdf_individual
  
  # model residuals
  E1_list[[i]] <- E1_individual
  
  # dispersion and dic
  dispersion_vector[i] <- Dispersion
  DIC_vector[i] <- I1$dic$dic
  
  # variograms
  v1_list[[i]] <- v1_individual
  v2_list[[i]] <- v2_individual
  
  # posterior marginal distribution for year random walk
  year_marg_list[[i]] <- year.marg[[32]]
  
  # percent change since survey start mean and 95% confidence interval
  percent_change_vector[[i]] <- Magdf$percent.change.mean[32]
  percent_change_intervals_vector[[i]] <- Magdf$percent.change.cri95[32]
  
}



### binding data

yearsm_bound <- bind_rows(yearsm_list)
dayY_bound <- bind_rows(dayY_list)
myData_bound <- bind_rows(myData_list)
magdf_bound <- bind_rows(magdf_list)
E1_bound <- bind_rows(E1_list)
v1_bound <- bind_rows(v1_list)
v2_bound <- bind_rows(v2_list)

additional_data <- data.frame(species = species_options,
                              dispersion = dispersion_vector,
                              DIC = DIC_vector,
                              mean = percent_change_vector,
                              interval = percent_change_intervals_vector)



### declutter

rm("bspecies",
   "DayY","dayY_individual","dayY_list",
   "E1","E1_individual","E1_list",
   "hyper.prec",
   "I1",
   "Magdf","magdf_individual","magdf_list",
   "MyData","myData_individual","myData_list",
   "tmarg",
   "V1","v1_individual","v1_list",
   "V2","v2_individual","v2_list",
   "year_marg_list","year.marg",
   "Yearsm","yearsm_individual","yearsm_list")

```

```{r sample percent change}
#| echo: false
#| fig.cap: "Mean percent change in population size since 1989. This figure is included in the published paper. The presence, absence, or strength of the trend visualized here was a main factor in criteria determination."
ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance") +
  geom_line(data = magdf_bound, aes(x = year, y = change), color = "black") +
  geom_ribbon(data = magdf_bound, aes(x = year, ymin = lower, ymax = upper), alpha = 0.3) +
  xlim(1989,2020) +
  facet_wrap(~ species,scales = "free") + 
  ggtitle("Percent change in Abundance Since 1989") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r sample variogram}
#| echo: false
#| fig.cap: "Variogram visualization of autocorrelation between years. Based on the sample study species, the existence of mild autocorrelation did not seem to impede model implementation."
ggplot(data = v1_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species) + 
  labs(x = "Time Difference (Years)",
       y = "Sample Variogram",
       title = "Variogram 1") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r sample abundance comparison}
#| echo: false
#| fig.cap: "Comparison of INLA model fitted abundance and recorded population abundance by day of the year. In many cases with CCBER data, the fitted and actual abundance values were very similar. The deviation of one from the other seems a little concerning, but not enough to break a model."
ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = fitted),color = "blue") + 
  geom_point(aes(x = day_of_year ,y = abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = day_of_year ,y = fitted),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species) + 
  labs(title = "Overlay of Fitted Abundance vs Day of Year (Blue) 
       and Abundance vs Day of Year (Red)") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r sample statistics}
#| echo: false
#| fig.cap: "A table of all meaningful statistics. The dispersion statistic was the most important factor in determining model fit. DIC tended to trend identically to dispersion, but with larger numbers. "
kable(additional_data)
```

```{r plotting replicated models and measures of fit}
#| include: false

### trend by year
ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = yearsm_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = yearsm_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ species) + 
  theme_bw() +
  ggtitle('Trend by Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(yearsm_bound$quant0.025)-0.05,max(yearsm_bound$quant0.975)+0.05)


### trend by day of year
ggplot() +
  xlab("Day") + ylab("Trend") +
  geom_line(data = dayY_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = dayY_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ species) + 
  theme_bw() +
  ggtitle('Trend by Day of Survey Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) + 
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5))


### variograms
# Variogram checking for temporal autocorrelation by year 
ggplot(data = v1_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species) + 
  labs(x = "Time Difference (Years)",
       y = "Sample Variogram",
       title = "Variogram 1") + 
  theme(plot.title = element_text(hjust = 0.5))

# Variogram checking for temporal autocorrelation by date of year
ggplot(data = v2_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species) + 
  labs(x = "Time Difference (Day of Year)",
       y = "Sample Variogram",
       title = "Variogram 2") + 
  theme(plot.title = element_text(hjust = 0.5))


### checking residuals

ggplot(data = E1_bound,
       aes(x = residuals,y = myData_bound$fitted)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  labs(title = "Fitted Abundance vs Residuals") + 
  theme(plot.title = element_text(hjust = 0.5))
# looking for adherence to 0 line with minor variability

ggplot(data = E1_bound,
       aes(sample = residuals)) + 
  geom_qq() + 
  geom_qq_line() + 
  facet_wrap(~ species) + 
  labs(title = "QQPlot") + 
  theme(plot.title = element_text(hjust = 0.5))
# looking for points within theoretical line bounds
# determines whether or not residuals are distributed normally

ggplot(E1_bound,
        aes(x = residuals)) + 
  geom_histogram() + 
  facet_wrap(~ species) + 
  labs(title = "Residuals Histogram") + 
  theme(plot.title = element_text(hjust = 0.5))


ggplot(data = myData_bound,
       aes(x = abundance, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  labs(title = "Residuals vs Species Abundance") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound,
       aes(x = abundance, y = fitted)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  labs(title = "Fitted Abundance vs Species Abundance") + 
  theme(plot.title = element_text(hjust = 0.5))


### residuals by time step
# looking for residual patterns

ggplot(myData_bound,
       aes(x = year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  labs(title = "Residuals vs Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(myData_bound,
       aes(x = day_of_year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  labs(title = "Residuals vs Day of Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = residuals)) +
  geom_smooth(aes(x = date ,y = residuals),
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  facet_wrap(~ species) + 
  labs(title = "Residuals vs Date") + 
  theme(plot.title = element_text(hjust = 0.5))


### comparing recorded abundance to fitted abundance

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance),color = "blue") + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_point(aes(x = date,y = fitted),color = "orange") + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  facet_wrap(~ species) + 
  labs(title = "Overlay of Abundance vs Date (Blue) and Fitted Abundance vs Date (Orange)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted), colour = 'orange', alpha = 1/10) +
  geom_point(aes(x = date,y = abundance), colour = 'blue', alpha = 1/10) +
  ylim(c(0,20)) + 
  facet_wrap(~ species) + 
  labs(title = "Overlay of Abundance vs Date (Blue) and Fitted Abundance vs Date (Orange)") + 
  theme(plot.title = element_text(hjust = 0.5))


### abundances by time measurements

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance)) + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species) + 
  labs(title = "Species Abundance vs Date") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted)) + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Fitted Abundance vs Date') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted Abundance') + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = abundance),color = "blue") + 
  geom_smooth(aes(x = day_of_year,y = abundance),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species) + 
  labs(title = "Abundance vs Day of Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = fitted),color = "blue") + 
  geom_point(aes(x = day_of_year ,y = abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = day_of_year ,y = fitted),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species) + 
  labs(title = "Overlay of Fitted Abundance vs Day of Year (Blue) 
       and Abundance vs Day of Year (Red)") + 
  theme(plot.title = element_text(hjust = 0.5))


### Percent Change in Abundance
# Plot percent change since 2017 and 95% CRIs for all years
ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance") +
  geom_line(data = magdf_bound, aes(x = year, y = change), color = "black") +
  geom_ribbon(data = magdf_bound, aes(x = year, ymin = lower, ymax = upper), alpha = 0.3) +
  xlim(1989,2020) +
  facet_wrap(~ species,scales = "free") + 
  ggtitle("Percent change in Abundance Since 2017") + 
  theme(plot.title = element_text(hjust = 0.5))

(additional_data)

```

Using these visualizations and statistics and another modification of the Julien-Tyler script, through trial and error, ideal U values and population distributions were determined for each of the nine example CCBER species: Common Yellowthroat, Killdeer, American Crow, Song Sparrow, Northern Shoveler, Least Sandpiper, Gadwall, Semipalmated Plover and Red-tailed Hawk. I took notes on the "diagnostic plots" output in a Google document and have comments to accompany my choices for U values and distributions.

```{r fitting models with ccber data, one species with a range of U values}
#| message: false
#| include: false

### U values range
# sample first test: c(0.0005, 0.005, 0.015, 0.05, 0.15, 0.5)
# c(0.5, 0.6, 0.7, 0.8, 0.9, 1)
# c(0.1, 0.15, 0.2, 0.25, 0.3, 0.4)
u_options <- c(0.05,0.1,0.15,0.2,0.3,0.4)



### initialization

# blank stuff
yearsm_list <- list()
dayY_list <- list()
myData_list <- list()
magdf_list <- list()
E1_list <- list()
dispersion_vector <- vector()
DIC_vector <- vector()
v1_list <- list()
v2_list <- list()
year_marg_list <- list()
percent_change_vector <- vector()
percent_change_intervals_vector <- vector()

# variograms
breaks <- seq(0.5, 20.5, length = 20)
breaks2 <- seq(0.5, 364.5, length = 365)



### running and saving INLA models with a default of U = 0.05

for (i in 1:length(u_options)){
  
  ### INLA model
  
    species <- "Semipalmated Plover"
    U <- u_options[i]
    distribution <- "nbinomial"

    speciescol <- species
    MyData <- na.exclude(data[, c(speciescol,
                                  "day_of_survey_year",
                                  "observation_date",
                                  'survey_year',
                                  'survey_id')])
    names(MyData)[1] <- 'Abundance'
    bspecies <- MyData[,1]

    hyper.prec <- list(theta = list(prior = "pc.prec",
                                    param = c(U, 0.05)))
    inla.setOption(scale.model.default = TRUE)

    I1 <- inla(Abundance ~ f(survey_year, model = "rw1", hyper = hyper.prec) + 
             f(day_of_survey_year, model = "rw2", hyper = hyper.prec),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)
    
  ### calculations and saving data
    
    MyData$Fitted <- I1$summary.fitted.values[,'mean']
    
    # trend for year
    Yearsm <- I1$summary.random$survey_year
    
    # calculating residuals
    mu <- MyData$Fitted
    theta <- I1$summary.hyperpar[1, "mean"]
    
    if(distribution == 'nbinomial'){
      # Pearson residuals for negative binomial distribution
      E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
    }
    if(distribution == 'poisson'){
      # Pearson residuals for poisson distribution
      E1 <- (bspecies - mu) / sqrt(mu) 
    }
    if(distribution == 'gaussian'){
      # Residuals for gaussian distribution
      E1 <-  MyData$Abundance-MyData$Fitted
    }
    names(E1)[1] <- "Residuals"
    
    
    ### Check for overdispersion
    
    N <- nrow(MyData)
    p <- nrow(I1$summary.fixed)
    Dispersion <- sum(E1^2) / (N - p)
    
    
    ### Day of Year Trends
    
    DayY <- I1$summary.random$day_of_survey_year
    MyData$Residuals <- E1$Residuals
    
    
    ### variograms
    
    # Variogram for checking for temporal autocorrelation by year 
    temp <- E1$Residuals
    MyData$E1 <- temp
    MyData$Zeros <- rep(0, nrow(MyData))
    V1 <- variog(coords = MyData[, c("survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks)
    
    # Variogram for checking for temporal autocorrelation by data of year
    V2 <- variog(coords = MyData[, c("day_of_survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks2)
    
    
    ### Extract the posterior marginal distributions for Year random walk
    survey_year.marg <- I1$marginals.random$survey_year 
    
    
    ### Create dataframe for storing data
    Magdf <- data.frame(survey_year = seq(from = 1, to = 8, by = 1))
    for(j in 1:8) { 
      # Transform the posterior marginal distribution for current year so that it is not in log scale
      tmarg <- inla.tmarginal(exp, survey_year.marg[[j]]) 
      # Calculate mean value for year i
      Magdf$mean[j] <- mean(tmarg[,1]) 
      # Calculate 95% CRI for year i's estimate
      Magdf$cri95[j] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                             inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
      # Calculate mean percent change current year vs initial
      Magdf$percent.change.mean[j] <- 100 * (Magdf$mean[j] - Magdf$mean[1]) / Magdf$mean[1] 
      # Calculate variance in percent change using 95% CRIs and variance estimation method 
      Magdf$percent.change.cri95[j] <- Magdf$mean[j] / Magdf$mean[1] *
        sqrt((Magdf$cri95[j]^2 / Magdf$mean[j]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
      Magdf$lower.cri95[j] <- Magdf$percent.change.mean[j] - (Magdf$percent.change.cri95[j])
      Magdf$upper.cri95[j] <- Magdf$percent.change.mean[j] + (Magdf$percent.change.cri95[j])
    }
    
    
    ### saving data
    
    yearsm_individual <- data.frame(u_value = u_options[i],
                                    ID = Yearsm$ID,
                                    mean = Yearsm$mean,
                                    quant0.025 = Yearsm$`0.025quant`,
                                    quant0.975 = Yearsm$`0.975quant`)
    
    dayY_individual <- data.frame(u_value = u_options[i],
                                  ID = DayY$ID,
                                  mean = DayY$mean,
                                  quant0.025 = DayY$`0.025quant`,
                                  quant0.975 = DayY$`0.975quant`)
    
    myData_individual <- data.frame(u_value = u_options[i],
                                    abundance = MyData$Abundance,
                                    day_of_year = MyData$day_of_survey_year,
                                    date = MyData$observation_date,
                                    survey_year = MyData$survey_year,
                                    fitted = MyData$Fitted,
                                    residuals = MyData$Residuals,
                                    E1 = MyData$E1)
    
    magdf_individual <- data.frame(u_value = u_options[i],
                                   survey_year = Magdf$survey_year,
                                   change = Magdf$percent.change.mean,
                                   lower = Magdf$lower.cri95,
                                   upper = Magdf$upper.cri95)
    
    E1_individual <- data.frame(u_value = u_options[i],
                                residuals = E1$Residuals)
    
    v1_individual <- data.frame(u_value = u_options[i],
                                u = V1$u,
                                v = V1$v)
    
    v2_individual <- data.frame(u_value = u_options[i],
                                u = V2$u,
                                v = V2$v)
    
  # data frames
  yearsm_list[[i]] <- yearsm_individual
  dayY_list[[i]] <- dayY_individual
  myData_list[[i]] <- myData_individual
  magdf_list[[i]] <- magdf_individual
  
  # model residuals
  E1_list[[i]] <- E1_individual
  
  # dispersion and dic
  dispersion_vector[i] <- Dispersion
  DIC_vector[i] <- I1$dic$dic
  
  # variograms
  v1_list[[i]] <- v1_individual
  v2_list[[i]] <- v2_individual
  
  # posterior marginal distribution for year random walk
  year_marg_list[[i]] <- survey_year.marg[[8]]
  
  # percent change since survey start mean and 95% confidence interval
  percent_change_vector[[i]] <- Magdf$percent.change.mean[8]
  percent_change_intervals_vector[[i]] <- Magdf$percent.change.cri95[8]
  
}



### binding data

yearsm_bound <- bind_rows(yearsm_list)
dayY_bound <- bind_rows(dayY_list)
myData_bound <- bind_rows(myData_list)
magdf_bound <- bind_rows(magdf_list)
E1_bound <- bind_rows(E1_list)
v1_bound <- bind_rows(v1_list)
v2_bound <- bind_rows(v2_list)

additional_data <- data.frame(u_value = u_options,
                              dispersion = dispersion_vector,
                              DIC = DIC_vector,
                              mean = percent_change_vector,
                              interval = percent_change_intervals_vector)



### declutter

rm("bspecies",
   "DayY","dayY_individual","dayY_list",
   "E1","E1_individual","E1_list",
   "hyper.prec",
   "I1",
   "Magdf","magdf_individual","magdf_list",
   "MyData","myData_individual","myData_list",
   "tmarg",
   "V1","v1_individual","v1_list",
   "V2","v2_individual","v2_list",
   "year_marg_list","survey_year.marg",
   "Yearsm","yearsm_individual","yearsm_list")



### plotting

### trend by year
ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = yearsm_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = yearsm_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ u_value) + 
  theme_bw() +
  ggtitle('Trend by Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(yearsm_bound$quant0.025)-0.05,max(yearsm_bound$quant0.975)+0.05)


### trend by day of year
ggplot() +
  xlab("Day") + ylab("Trend") +
  geom_line(data = dayY_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = dayY_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ u_value) + 
  theme_bw() +
  ggtitle('Trend by Day of Survey Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) + 
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5))


### variograms
# Variogram checking for temporal autocorrelation by year 
ggplot(data = v1_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ u_value) + 
  labs(x = "Time Difference (Years)",
       y = "Sample Variogram",
       title = "Variogram 1") + 
  theme(plot.title = element_text(hjust = 0.5))

# Variogram checking for temporal autocorrelation by date of year
ggplot(data = v2_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ u_value) + 
  labs(x = "Time Difference (Day of Year)",
       y = "Sample Variogram",
       title = "Variogram 2") + 
  theme(plot.title = element_text(hjust = 0.5))


### checking residuals

ggplot(data = E1_bound,
       aes(x = residuals,y = myData_bound$fitted)) + 
  geom_point() + 
  facet_wrap(~ u_value) + 
  labs(title = "Fitted Abundance vs Residuals") + 
  theme(plot.title = element_text(hjust = 0.5))
# looking for adherence to 0 line with minor variability

ggplot(data = E1_bound,
       aes(sample = residuals)) + 
  geom_qq() + 
  geom_qq_line() + 
  facet_wrap(~ u_value) + 
  labs(title = "QQPlot") + 
  theme(plot.title = element_text(hjust = 0.5))
# looking for points within theoretical line bounds
# determines whether or not residuals are distributed normally

ggplot(E1_bound,
        aes(x = residuals)) + 
  geom_histogram() + 
  facet_wrap(~ u_value) + 
  labs(title = "Residuals Histogram") + 
  theme(plot.title = element_text(hjust = 0.5))


ggplot(data = myData_bound,
       aes(x = abundance, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ u_value) + 
  labs(title = "Residuals vs Species Abundance") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound,
       aes(x = abundance, y = fitted)) + 
  geom_point() + 
  facet_wrap(~ u_value) + 
  labs(title = "Fitted Abundance vs Species Abundance") + 
  theme(plot.title = element_text(hjust = 0.5))


### residuals by time step
# looking for residual patterns

ggplot(myData_bound,
       aes(x = year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ u_value) + 
  labs(title = "Residuals vs Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(myData_bound,
       aes(x = day_of_year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ u_value) + 
  labs(title = "Residuals vs Day of Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = residuals)) +
  geom_smooth(aes(x = date ,y = residuals),
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  facet_wrap(~ u_value) + 
  labs(title = "Residuals vs Date") + 
  theme(plot.title = element_text(hjust = 0.5))


### comparing recorded abundance to fitted abundance

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance),color = "blue") + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_point(aes(x = date,y = fitted),color = "orange") + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  facet_wrap(~ u_value) + 
  labs(title = "Overlay of Abundance vs Date (Blue) and Fitted Abundance vs Date (Orange)") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted), colour = 'orange', alpha = 5/10) +
  geom_point(aes(x = date,y = abundance), colour = 'blue', alpha = 5/10) +
  facet_wrap(~ u_value) + 
  labs(title = "Overlay of Abundance vs Date (Blue) and Fitted Abundance vs Date (Orange)") + 
  theme(plot.title = element_text(hjust = 0.5))


### abundances by time measurements

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance)) + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ u_value) + 
  labs(title = "Species Abundance vs Date") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted)) + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Fitted Abundance vs Date') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted Abundance') + 
  facet_wrap(~ u_value)

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = abundance),color = "blue") + 
  geom_smooth(aes(x = day_of_year,y = abundance),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ u_value) + 
  labs(title = "Abundance vs Day of Year") + 
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = fitted),color = "blue") + 
  geom_point(aes(x = day_of_year ,y = abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = day_of_year ,y = fitted),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ u_value) + 
  labs(title = "Overlay of Fitted Abundance vs Day of Year (Blue) 
       and Abundance vs Day of Year (Red)") + 
  theme(plot.title = element_text(hjust = 0.5))


### Percent Change in Abundance
# Plot percent change since 2017 and 95% CRIs for all years
ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance") +
  geom_line(data = magdf_bound, aes(x = year, y = change), color = "black") +
  geom_ribbon(data = magdf_bound, aes(x = year, ymin = lower, ymax = upper), alpha = 0.3) +
  xlim(2017,2025) +
  facet_wrap(~ u_value, scales = "free_y") + 
  ggtitle("Percent change in Abundance Since 2017") + 
  theme(plot.title = element_text(hjust = 0.5))

(additional_data)

```

```{r fitting models with ccber data, multiple sample species}
#| message: false
#| include: false

### setting desired NCOS species

species_options <- c("Common Yellowthroat",
                     "Killdeer",
                     "American Crow",
                     "Song Sparrow",
                     "Northern Shoveler",
                     "Least Sandpiper",
                     "Gadwall",
                     "Semipalmated Plover",
                     "Red-shouldered Hawk")



### initialization

yearsm_list <- list()
dayY_list <- list()
myData_list <- list()
magdf_list <- list()
E1_list <- list()
dispersion_vector <- vector()
DIC_vector <- vector()
v1_list <- list()
v2_list <- list()
year_marg_list <- list()
percent_change_vector <- vector()
percent_change_intervals_vector <- vector()

breaks <- seq(0.5, 20.5, length = 20)
breaks2 <- seq(0.5, 364.5, length = 365)



### running and saving INLA models with a default of U = 0.05

for (i in 1:length(species_options)){
  
  ### INLA model
  
    species <- species_options[i]
    
    if (species == "Common Yellowthroat"){U <- 1}
    if (species == "Killdeer" | species == "Song Sparrow"){U <- 0.5}
    if (species == "American Crow"){U <- 0.6}
    if (species == "Northern Shoveler"){U <- 0.05}
    if (species == "Least Sandpiper"){U <- 0.7}
    if (species == "Gadwall" | species == "Semipalmated Plover"){U <- 0.3}
    if (species == "Red-tailed Hawk"){U <- 0.01}
    
    if (species == "Common Yellowthroat"){distribution <- "poisson"}
    else {distribution <- "nbinomial"}

    speciescol <- species
    MyData <- na.exclude(data[, c(speciescol,
                                  "day_of_survey_year",
                                  "observation_date",
                                  'survey_year',
                                  'survey_id')])
    names(MyData)[1] <- 'Abundance'
    bspecies <- MyData[,1]

    hyper.prec <- list(theta = list(prior = "pc.prec",
                                    param = c(U, 0.05)))
    inla.setOption(scale.model.default = TRUE)

    I1 <- inla(Abundance ~ f(survey_year, model = "rw1", hyper = hyper.prec) + 
             f(day_of_survey_year, model = "rw2", hyper = hyper.prec),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)
    
  ### calculations and saving data
    
    MyData$Fitted <- I1$summary.fitted.values[,'mean']
    
    # trend for year
    Yearsm <- I1$summary.random$survey_year
    
    # calculating residuals
    mu <- MyData$Fitted
    theta <- I1$summary.hyperpar[1, "mean"]
    
    if(distribution == 'nbinomial'){
      # Pearson residuals for negative binomial distribution
      E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
    }
    if(distribution == 'poisson'){
      # Pearson residuals for poisson distribution
      E1 <- (bspecies - mu) / sqrt(mu) 
    }
    if(distribution == 'gaussian'){
      # Residuals for gaussian distribution
      E1 <-  MyData$Abundance-MyData$Fitted
    }
    names(E1)[1] <- "Residuals"
    
    
    ### Check for overdispersion
    
    N <- nrow(MyData)
    p <- nrow(I1$summary.fixed)
    Dispersion <- sum(E1^2) / (N - p)
    
    
    ### Day of Year Trends
    
    DayY <- I1$summary.random$day_of_survey_year
    MyData$Residuals <- E1$Residuals
    
    
    ### variograms
    
    # Variogram for checking for temporal autocorrelation by year 
    temp <- E1$Residuals
    MyData$E1 <- temp
    MyData$Zeros <- rep(0, nrow(MyData))
    V1 <- variog(coords = MyData[, c("survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks)
    
    # Variogram for checking for temporal autocorrelation by data of year
    V2 <- variog(coords = MyData[, c("day_of_survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks2)
    
    
    ### Extract the posterior marginal distributions for Year random walk
    survey_year.marg <- I1$marginals.random$survey_year 
    
    
    ### Create dataframe for storing data
    Magdf <- data.frame(survey_year = seq(from = 1, to = 8, by = 1))
    for(j in 1:8) { 
      # Transform the posterior marginal distribution for current year so that it is not in log scale
      tmarg <- inla.tmarginal(exp, survey_year.marg[[j]]) 
      # Calculate mean value for year i
      Magdf$mean[j] <- mean(tmarg[,1]) 
      # Calculate 95% CRI for year i's estimate
      Magdf$cri95[j] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                             inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
      # Calculate mean percent change current year vs initial
      Magdf$percent.change.mean[j] <- 100 * (Magdf$mean[j] - Magdf$mean[1]) / Magdf$mean[1] 
      # Calculate variance in percent change using 95% CRIs and variance estimation method 
      Magdf$percent.change.cri95[j] <- Magdf$mean[j] / Magdf$mean[1] *
        sqrt((Magdf$cri95[j]^2 / Magdf$mean[j]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
      Magdf$lower.cri95[j] <- Magdf$percent.change.mean[j] - (Magdf$percent.change.cri95[j])
      Magdf$upper.cri95[j] <- Magdf$percent.change.mean[j] + (Magdf$percent.change.cri95[j])
    }
    
    
    ### saving data
    
    yearsm_individual <- data.frame(species = species_options[i],
                                    ID = Yearsm$ID,
                                    mean = Yearsm$mean,
                                    quant0.025 = Yearsm$`0.025quant`,
                                    quant0.975 = Yearsm$`0.975quant`)
    
    dayY_individual <- data.frame(species = species_options[i],
                                  ID = DayY$ID,
                                  mean = DayY$mean,
                                  quant0.025 = DayY$`0.025quant`,
                                  quant0.975 = DayY$`0.975quant`)
    
    myData_individual <- data.frame(species = species_options[i],
                                    abundance = MyData$Abundance,
                                    day_of_year = MyData$day_of_survey_year,
                                    date = MyData$observation_date,
                                    survey_year = MyData$survey_year,
                                    fitted = MyData$Fitted,
                                    residuals = MyData$Residuals,
                                    E1 = MyData$E1)
    
    magdf_individual <- data.frame(species = species_options[i],
                                   survey_year = Magdf$survey_year,
                                   change = Magdf$percent.change.mean,
                                   lower = Magdf$lower.cri95,
                                   upper = Magdf$upper.cri95)
    
    E1_individual <- data.frame(species = species_options[i],
                                residuals = E1$Residuals)
    
    v1_individual <- data.frame(species = species_options[i],
                                u = V1$u,
                                v = V1$v)
    
    v2_individual <- data.frame(species = species_options[i],
                                u = V2$u,
                                v = V2$v)
    
  # data frames
  yearsm_list[[i]] <- yearsm_individual
  dayY_list[[i]] <- dayY_individual
  myData_list[[i]] <- myData_individual
  magdf_list[[i]] <- magdf_individual
  
  # model residuals
  E1_list[[i]] <- E1_individual
  
  # dispersion and dic
  dispersion_vector[i] <- Dispersion
  DIC_vector[i] <- I1$dic$dic
  
  # variograms
  v1_list[[i]] <- v1_individual
  v2_list[[i]] <- v2_individual
  
  # posterior marginal distribution for year random walk
  year_marg_list[[i]] <- survey_year.marg[[8]]
  
  # percent change since survey start mean and 95% confidence interval
  percent_change_vector[[i]] <- Magdf$percent.change.mean[8]
  percent_change_intervals_vector[[i]] <- Magdf$percent.change.cri95[8]
  
}



### binding data

yearsm_bound <- bind_rows(yearsm_list)
dayY_bound <- bind_rows(dayY_list)
myData_bound <- bind_rows(myData_list)
magdf_bound <- bind_rows(magdf_list)
E1_bound <- bind_rows(E1_list)
v1_bound <- bind_rows(v1_list)
v2_bound <- bind_rows(v2_list)

additional_data <- data.frame(species = species_options,
                              dispersion = dispersion_vector,
                              DIC = DIC_vector,
                              mean = percent_change_vector,
                              interval = percent_change_intervals_vector)



### declutter

rm("bspecies",
   "DayY","dayY_individual","dayY_list",
   "E1","E1_individual","E1_list",
   "hyper.prec",
   "I1",
   "Magdf","magdf_individual","magdf_list",
   "MyData","myData_individual","myData_list",
   "tmarg",
   "V1","v1_individual","v1_list",
   "V2","v2_individual","v2_list",
   "year_marg_list","survey_year.marg",
   "Yearsm","yearsm_individual","yearsm_list")

```

```{r example species percent change}
#| echo: false
plot1 <- ggplot() +
  xlab("Survey Year") +
  ylab("Percent Change in Abundance since 2017") +
  ggtitle("Percent Change in Abundance since 2017 versus Survey Year by Species") + 
  geom_line(data = magdf_bound, aes(x = survey_year, y = change), color = "black") +
  geom_ribbon(data = magdf_bound, aes(x = survey_year, ymin = lower, ymax = upper), alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") + 
  facet_wrap(~ species,scales = "free_y") + 
  scale_x_continuous(breaks = seq(1, 8, by = 1)) + 
  theme_light() + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 13))

ggsave(filename = paste0("Sample_CCBER_Species_PCA_",
                         format(Sys.Date(), "%y-%m-%d"), ".pdf"),
       plot1,
       width = 6.67,
       height = 5,
       units = "in")

```

```{r example species variogram}
#| echo: false
ggplot(data = v1_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species, scales = "free_y") + 
  labs(x = "Time Difference (Survey Years)",
       y = "Sample Variogram")
```

```{r example species abundance comparison}
#| echo: false
ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = fitted),color = "blue") + 
  geom_point(aes(x = day_of_year ,y = abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = day_of_year ,y = fitted),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species, scales = "free_y") + 
  ggtitle("Overlay of Fitted Abundance vs Day of Year (Blue) 
       and Abundance vs Day of Year (Red)") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r example species statistics}
#| echo: false
kable(additional_data)
```

```{r plotting ccber species inla models}
#| include: false

### trend by year
ggplot() +
  xlab("Year") + ylab("Trend") +
  geom_line(data = yearsm_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = yearsm_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ species) + 
  theme_bw() +
  ggtitle('Trend by Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) +
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5)) +
  # ylim(-0.5,0.5) +
  ylim(min(yearsm_bound$quant0.025)-0.05,max(yearsm_bound$quant0.975)+0.05)


### trend by day of year
ggplot() +
  xlab("Day") + ylab("Trend") +
  geom_line(data = dayY_bound, aes(x = ID, y = mean)) +
  geom_ribbon(data = dayY_bound, aes(x = ID, ymax = quant0.025, ymin = quant0.975), alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~ species) + 
  theme_bw() +
  ggtitle('Trend by Day of Survey Year') +
  theme(axis.text.y = element_text(size = 8)) +
  theme(axis.text.x = element_text(size = 8)) +
  theme(axis.title = element_text(size = 11)) + 
  theme(plot.title = element_text(size = 15,
                                  lineheight=.8, 
                                  face="bold", 
                                  hjust = 0.5))


### residuals by date
ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = residuals)) +
  geom_smooth(aes(x = date ,y = residuals),
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  facet_wrap(~ species)
# looking for residual patterns


### Variogram checking for temporal autocorrelation by year 
ggplot(data = v1_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species) + 
  labs(x = "Time Difference (Years)",
       y = "Sample Variogram")

### Variogram checking for temporal autocorrelation by date of year
ggplot(data = v2_bound,aes(x = u,y = v)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~ species) + 
  labs(x = "Time Difference (Day of Year)",
       y = "Sample Variogram")


### Plot Residuals vs. fitted values and covariates

ggplot(data = E1_bound,
       aes(x = residuals,y = myData_bound$fitted)) + 
  geom_point() + 
  facet_wrap(~ species)
# looking for adherence to 0 line with minor variability

ggplot(data = E1_bound,
       aes(sample = residuals)) + 
  geom_qq() + 
  geom_qq_line() + 
  facet_wrap(~ species)
# looking for points within theoretical line bounds
# determines whether or not residuals are distributed normally

ggplot(E1_bound,
        aes(x = residuals)) + 
  geom_histogram() + 
  facet_wrap(~ species)

ggplot(data = myData_bound,
       aes(x = abundance, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species)

ggplot(data = myData_bound,
       aes(x = abundance, y = fitted)) + 
  geom_point() + 
  facet_wrap(~ species)


### residuals by year or day of year

ggplot(myData_bound,
       aes(x = survey_year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species)

ggplot(myData_bound,
       aes(x = day_of_year, y = E1_bound$residuals)) + 
  geom_point() + 
  facet_wrap(~ species)


### comparing recorded abundance to model fitted values
# adding geom_point to the appropriate ggplots below since geom_smooth is still not working

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance)) + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = abundance),color = "blue") + 
  geom_smooth(aes(x = date ,y = abundance), 
              span = 0.2, colour = 'blue',  
              method = 'loess') +
  geom_point(aes(x = date,y = fitted),color = "orange") + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted)) + 
  geom_smooth(aes(x = date ,y = fitted), 
              span = 0.2, colour = 'black',  
              method = 'loess') + 
  ggtitle('Species Abundance') +
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('Fitted abundance') + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = date,y = fitted), colour = 'blue', alpha = 1/10) +
  geom_point(aes(x = date,y = abundance), colour = 'red', alpha = 1/10) +
  ylim(c(0,20)) + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = abundance),color = "blue") + 
  geom_smooth(aes(x = day_of_year,y = abundance),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species)

ggplot(data = myData_bound) +
  geom_point(aes(x = day_of_year,y = fitted),color = "blue") + 
  geom_point(aes(x = day_of_year ,y = abundance), colour = 'red', alpha = 1/10) +
  geom_smooth(aes(x = day_of_year ,y = fitted),
              span = 0.2, colour = 'blue',  
              method = 'loess') + 
  facet_wrap(~ species)


### Percent Change in Abundance
# Plot percent change since 2017 and 95% CRIs for all years
ggplot() +
  xlab("Year") +
  ylab("Percent change in abundance since 1989") +
  geom_line(data = magdf_bound, aes(x = survey_year, y = change), color = "black") +
  geom_ribbon(data = magdf_bound, aes(x = survey_year, ymin = lower, ymax = upper), alpha = 0.3) +
  xlim(1,8) +
  facet_wrap(~ species,
             scales = "free_y")

```

```{r abundance vs survey year}

abundance_year <- ggplot(myData_bound,
       aes(x = survey_year,
           y = abundance)) + 
  geom_point() + 
  facet_wrap(~ species) + 
  theme_light() + 
  labs(title = "Sample CCBER Species Abundance versus Survey Year",
       x = "Survey Year",
       y = "Abundance") + 
  theme(plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 15))

ggsave(filename = paste0("Sample_CCBER_Species_Abundance_",
                         format(Sys.Date(), "%y-%m-%d"), ".pdf"),
       abundance_year,
       width = 200,
       height = 140,
       units = "mm")

```

```{r fitting models with ccber data, at least 1/3 frequency species}

# obtaining all species observed in greater than half (48) of surveys (i.e., 49 or more)

presence_path <- here("data","Matt_Vinh","compiled_and_cleaned_2026-02-18.rds")
observation_level <- read_rds(presence_path)

species_presence <- observation_level %>% 
  filter(category == "species") %>% 
  group_by(species) %>% 
  summarise(num_surveys = n_distinct(observation_date)) %>% 
  arrange(desc(num_surveys))
over_half <- species_presence %>% 
  filter(num_surveys >= 33)



### initialization

yearsm_list <- list()
dayY_list <- list()
myData_list <- list()
magdf_list <- list()
E1_list <- list()
dispersion_vector <- vector()
DIC_vector <- vector()
v1_list <- list()
v2_list <- list()
year_marg_list <- list()
percent_change_vector <- vector()
percent_change_intervals_vector <- vector()

breaks <- seq(0.5, 20.5, length = 20)
breaks2 <- seq(0.5, 364.5, length = 365)



### inla models

for (i in 1:length(over_half$species)){
  
  ### INLA model
  
    species <- over_half$species[i]
    
    if (species == "Common Yellowthroat"){U <- 1}
    if (species == "Killdeer" | species == "Song Sparrow"){U <- 0.5}
    if (species == "American Crow"){U <- 0.6}
    if (species == "Northern Shoveler"){U <- 0.05}
    if (species == "Least Sandpiper"){U <- 0.7}
    if (species == "Gadwall" | species == "Semipalmated Plover"){U <- 0.3}
    if (species == "Red-tailed Hawk"){U <- 0.01}
    else {U <- 0.04} # median U value from julien-tyler paper
    
    if (species %in% c("Killdeer", "Song Sparrow", "American Crow", "Northern Shoveler",
                       "Least Sandpiper", "Gadwall", "Red-tailed Hawk"))
      {distribution <- "nbinomial"}
    else {distribution <- "poisson"}

    speciescol <- species
    MyData <- na.exclude(data[, c(speciescol,
                                  "day_of_survey_year",
                                  "observation_date",
                                  'survey_year',
                                  'survey_id')])
    names(MyData)[1] <- 'Abundance'
    bspecies <- MyData[,1]

    hyper.prec <- list(theta = list(prior = "pc.prec",
                                    param = c(U, 0.05)))
    inla.setOption(scale.model.default = TRUE)

    I1 <- inla(Abundance ~ f(survey_year, model = "rw1", hyper = hyper.prec) + 
             f(day_of_survey_year, model = "rw2", hyper = hyper.prec),
           control.predictor = list(compute = TRUE, 
                                    quantiles = c("0.025", "0.975")),
           control.compute = list(dic = TRUE,
                                  waic = TRUE),
           family = distribution,
           data = MyData)
    
  ### calculations and saving data
    
    MyData$Fitted <- I1$summary.fitted.values[,'mean']
    
    # trend for year
    Yearsm <- I1$summary.random$survey_year
    
    # calculating residuals
    mu <- MyData$Fitted
    theta <- I1$summary.hyperpar[1, "mean"]
    
    if(distribution == 'nbinomial'){
      # Pearson residuals for negative binomial distribution
      E1 <- (bspecies - mu) / sqrt(mu + mu^2 / theta) 
    }
    if(distribution == 'poisson'){
      # Pearson residuals for poisson distribution
      E1 <- (bspecies - mu) / sqrt(mu) 
    }
    if(distribution == 'gaussian'){
      # Residuals for gaussian distribution
      E1 <-  MyData$Abundance-MyData$Fitted
    }
    names(E1)[1] <- "Residuals"
    
    
    ### Check for overdispersion
    
    N <- nrow(MyData)
    p <- nrow(I1$summary.fixed)
    Dispersion <- sum(E1^2) / (N - p)
    
    
    ### Day of Year Trends
    
    DayY <- I1$summary.random$day_of_survey_year
    MyData$Residuals <- E1$Residuals
    
    
    ### variograms
    
    # Variogram for checking for temporal autocorrelation by year 
    temp <- E1$Residuals
    MyData$E1 <- temp
    MyData$Zeros <- rep(0, nrow(MyData))
    V1 <- variog(coords = MyData[, c("survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks)
    
    # Variogram for checking for temporal autocorrelation by data of year
    V2 <- variog(coords = MyData[, c("day_of_survey_year", "Zeros")],
                 data = MyData$E1,
                 breaks = breaks2)
    
    
    ### Extract the posterior marginal distributions for Year random walk
    survey_year.marg <- I1$marginals.random$survey_year 
    
    
    ### Create dataframe for storing data
    Magdf <- data.frame(survey_year = seq(from = 1, to = 8, by = 1))
    for(j in 1:8) { 
      # Transform the posterior marginal distribution for current year so that it is not in log scale
      tmarg <- inla.tmarginal(exp, survey_year.marg[[j]]) 
      # Calculate mean value for year i
      Magdf$mean[j] <- mean(tmarg[,1]) 
      # Calculate 95% CRI for year i's estimate
      Magdf$cri95[j] <- mean(inla.qmarginal(0.5, tmarg) - inla.qmarginal(0.025, tmarg), 
                             inla.qmarginal(0.975, tmarg) - inla.qmarginal(0.5, tmarg))
      # Calculate mean percent change current year vs initial
      Magdf$percent.change.mean[j] <- 100 * (Magdf$mean[j] - Magdf$mean[1]) / Magdf$mean[1] 
      # Calculate variance in percent change using 95% CRIs and variance estimation method 
      Magdf$percent.change.cri95[j] <- Magdf$mean[j] / Magdf$mean[1] *
        sqrt((Magdf$cri95[j]^2 / Magdf$mean[j]^2) + (Magdf$cri95[1]^2 / Magdf$mean[1]^2)) * 100
      Magdf$lower.cri95[j] <- Magdf$percent.change.mean[j] - (Magdf$percent.change.cri95[j])
      Magdf$upper.cri95[j] <- Magdf$percent.change.mean[j] + (Magdf$percent.change.cri95[j])
    }
    
    
    ### saving data
    
    yearsm_individual <- data.frame(species = over_half$species[i],
                                    ID = Yearsm$ID,
                                    mean = Yearsm$mean,
                                    quant0.025 = Yearsm$`0.025quant`,
                                    quant0.975 = Yearsm$`0.975quant`)
    
    dayY_individual <- data.frame(species = over_half$species[i],
                                  ID = DayY$ID,
                                  mean = DayY$mean,
                                  quant0.025 = DayY$`0.025quant`,
                                  quant0.975 = DayY$`0.975quant`)
    
    myData_individual <- data.frame(species = over_half$species[i],
                                    abundance = MyData$Abundance,
                                    day_of_year = MyData$day_of_survey_year,
                                    date = MyData$observation_date,
                                    survey_year = MyData$survey_year,
                                    fitted = MyData$Fitted,
                                    residuals = MyData$Residuals,
                                    E1 = MyData$E1)
    
    magdf_individual <- data.frame(species = over_half$species[i],
                                   survey_year = Magdf$survey_year,
                                   change = Magdf$percent.change.mean,
                                   lower = Magdf$lower.cri95,
                                   upper = Magdf$upper.cri95)
    
    E1_individual <- data.frame(species = over_half$species[i],
                                residuals = E1$Residuals)
    
    v1_individual <- data.frame(species = over_half$species[i],
                                u = V1$u,
                                v = V1$v)
    
    v2_individual <- data.frame(species = over_half$species[i],
                                u = V2$u,
                                v = V2$v)
    
  # data frames
  yearsm_list[[i]] <- yearsm_individual
  dayY_list[[i]] <- dayY_individual
  myData_list[[i]] <- myData_individual
  magdf_list[[i]] <- magdf_individual
  
  # model residuals
  E1_list[[i]] <- E1_individual
  
  # dispersion and dic
  dispersion_vector[i] <- Dispersion
  DIC_vector[i] <- I1$dic$dic
  
  # variograms
  v1_list[[i]] <- v1_individual
  v2_list[[i]] <- v2_individual
  
  # posterior marginal distribution for year random walk
  year_marg_list[[i]] <- survey_year.marg[[8]]
  
  # percent change since survey start mean and 95% confidence interval
  percent_change_vector[[i]] <- Magdf$percent.change.mean[8]
  percent_change_intervals_vector[[i]] <- Magdf$percent.change.cri95[8]
  
}



### binding data

yearsm_bound <- bind_rows(yearsm_list)
dayY_bound <- bind_rows(dayY_list)
myData_bound <- bind_rows(myData_list)
magdf_bound <- bind_rows(magdf_list)
E1_bound <- bind_rows(E1_list)
v1_bound <- bind_rows(v1_list)
v2_bound <- bind_rows(v2_list)

additional_data <- data.frame(species = over_half$species,
                              dispersion = dispersion_vector,
                              DIC = DIC_vector,
                              mean = percent_change_vector,
                              interval = percent_change_intervals_vector)



### declutter

rm("bspecies",
   "DayY","dayY_individual","dayY_list",
   "E1","E1_individual","E1_list",
   "hyper.prec",
   "I1",
   "Magdf","magdf_individual","magdf_list",
   "MyData","myData_individual","myData_list",
   "tmarg",
   "V1","v1_individual","v1_list",
   "V2","v2_individual","v2_list",
   "year_marg_list","survey_year.marg",
   "Yearsm","yearsm_individual","yearsm_list")

```



## average annual trend visualization

```{r aat plot test}

# calculations

magdf_restricted <- magdf_bound %>% filter(year == 2020)

AAT <- data.frame(species = magdf_restricted$species,
                  PTM_Average_Annual_Trend = NA,
                  AAT_LowerCI = NA,
                  AAT_UpperCI = NA,
                  AAT_Credible_Interval = NA)

mag <- magdf_restricted$change * 0.01
AAT$PTM_Average_Annual_Trend <- ((1+mag)^(1/31)-1)*100 #PTM: population trend magnitude

mag <- magdf_restricted$lower * 0.01
AAT$AAT_LowerCI <- ((1+mag)^(1/31)-1)*100

mag <- magdf_restricted$upper * 0.01
AAT$AAT_UpperCI <- ((1+mag)^(1/31)-1)*100

AAT$AAT_Credible_Interval <- (abs(AAT$AAT_LowerCI - AAT$PTM_Average_Annual_Trend) +         
                              abs(AAT$AAT_UpperCI - AAT$PTM_Average_Annual_Trend))/2

# modification

limAAT <- AAT[!is.na(AAT$AAT_Credible_Interval), ]
limAAT <- limAAT[order(-limAAT$PTM_Average_Annual_Trend), ]
limAAT$species <- factor(limAAT$species,
                         levels = unique(limAAT$species))

# plotting

ggplot(limAAT, aes(x=species, y=PTM_Average_Annual_Trend)) + 
  geom_point(size=5) + 
  geom_segment(aes(x=species, 
                   xend=species, 
                   y=AAT_LowerCI, 
                   yend=AAT_UpperCI),
               linewidth = 1) +
  geom_hline(yintercept = c(0), linetype="dotted") +
  xlab("Species") +
  ylab("Annual Trend (1989-2020)") +
  theme(axis.title.y = element_text(size = 30, face = 'bold')) +
  theme(axis.title.x = element_text(size = 30, face = 'bold')) +
  theme(axis.text=element_text(size = 20, color = 'black')) +
  coord_flip()

# appears to be the correct implementation!

```

```{r jrbp replication test}
# using annual average instead of last year recorded

# calculations

magdf_mean <- magdf_bound %>% 
  group_by(species) %>% 
  summarise(
    mean_change = mean(change),
    mean_lower = mean(lower),
    mean_upper = mean(upper)
  )

AAT <- data.frame(species = magdf_mean$species,
                  PTM_Average_Annual_Trend = NA,
                  AAT_LowerCI = NA,
                  AAT_UpperCI = NA,
                  AAT_Credible_Interval = NA)

mag <- magdf_mean$mean_change * 0.01
AAT$PTM_Average_Annual_Trend <- ((1+mag)^(1/31)-1)*100 #PTM: population trend magnitude

mag <- magdf_mean$mean_lower * 0.01
AAT$AAT_LowerCI <- ((1+mag)^(1/31)-1)*100

mag <- magdf_mean$mean_upper * 0.01
AAT$AAT_UpperCI <- ((1+mag)^(1/31)-1)*100

AAT$AAT_Credible_Interval <- (abs(AAT$AAT_LowerCI - AAT$PTM_Average_Annual_Trend) +         
                              abs(AAT$AAT_UpperCI - AAT$PTM_Average_Annual_Trend))/2

# modification

limAAT <- AAT[!is.na(AAT$AAT_Credible_Interval), ]
limAAT <- limAAT[order(-limAAT$PTM_Average_Annual_Trend), ]
limAAT$species <- factor(limAAT$species,
                         levels = unique(limAAT$species))

# plotting

ggplot(limAAT, aes(x=species, y=PTM_Average_Annual_Trend)) + 
  geom_point(size=5) + 
  geom_segment(aes(x=species, 
                   xend=species, 
                   y=AAT_LowerCI, 
                   yend=AAT_UpperCI),
               linewidth = 1) +
  geom_hline(yintercept = c(0), linetype="dotted") +
  xlab("Species") +
  ylab("Annual Trend (1989-2020)") +
  theme(axis.title.y = element_text(size = 30, face = 'bold')) +
  theme(axis.title.x = element_text(size = 30, face = 'bold')) +
  theme(axis.text=element_text(size = 20, color = 'black')) +
  coord_flip()

# differes from AAT plot included in paper

```

```{r aat plot for ccber sample species}

# calculations

magdf_restricted <- magdf_bound %>% filter(survey_year == 8)

AAT <- data.frame(species = magdf_restricted$species,
                  PTM_Average_Annual_Trend = NA,
                  AAT_LowerCI = NA,
                  AAT_UpperCI = NA,
                  AAT_Credible_Interval = NA)

mag <- magdf_restricted$change * 0.01
AAT$PTM_Average_Annual_Trend <- ((1+mag)^(1/7)-1)*100 #PTM: population trend magnitude

mag <- magdf_restricted$lower * 0.01
AAT$AAT_LowerCI <- ((1+mag)^(1/7)-1)*100

mag <- magdf_restricted$upper * 0.01
AAT$AAT_UpperCI <- ((1+mag)^(1/7)-1)*100

AAT$AAT_Credible_Interval <- (abs(AAT$AAT_LowerCI - AAT$PTM_Average_Annual_Trend) +         
                              abs(AAT$AAT_UpperCI - AAT$PTM_Average_Annual_Trend))/2

# modification

limAAT <- AAT[!is.na(AAT$AAT_Credible_Interval), ]
limAAT <- limAAT[order(-limAAT$PTM_Average_Annual_Trend), ]
limAAT$species <- factor(limAAT$species,
                         levels = unique(limAAT$species))

# plotting

aat1 <- ggplot(limAAT, aes(x=species, y=PTM_Average_Annual_Trend)) + 
  geom_point(size=5) + 
  geom_segment(aes(x=species, 
                   xend=species, 
                   y=AAT_LowerCI, 
                   yend=AAT_UpperCI),
               linewidth = 1) +
  geom_hline(yintercept = c(0), linetype="dotted") +
  xlab("Species") +
  ylab("Annual Trend At Survey Year 8") +
  coord_flip() + 
  theme_classic()
# trends match the "percent change in abundance" plots

ggsave(filename = paste0("Sample_CCBER_Species_AAT_fixed_int_",
                         format(Sys.Date(), "%y-%m-%d"), ".pdf"),
       aat1,
       width = 13.33,
       height = 10,
       units = "in")

```

```{r aat plot for ccber at least 1/3 frequency species}

# calculations

magdf_restricted <- magdf_bound %>% filter(survey_year == 8)

AAT <- data.frame(species = magdf_restricted$species,
                  PTM_Average_Annual_Trend = NA,
                  AAT_LowerCI = NA,
                  AAT_UpperCI = NA,
                  AAT_Credible_Interval = NA)

mag <- magdf_restricted$change * 0.01
AAT$PTM_Average_Annual_Trend <- ((1+mag)^(1/8)-1)*100 #PTM: population trend magnitude

mag <- magdf_restricted$lower * 0.01
AAT$AAT_LowerCI <- ((1+mag)^(1/8)-1)*100

mag <- magdf_restricted$upper * 0.01
AAT$AAT_UpperCI <- ((1+mag)^(1/8)-1)*100

AAT$AAT_Credible_Interval <- (abs(AAT$AAT_LowerCI - AAT$PTM_Average_Annual_Trend) +         
                              abs(AAT$AAT_UpperCI - AAT$PTM_Average_Annual_Trend))/2

# modification

limAAT <- AAT[!is.na(AAT$AAT_Credible_Interval), ]
limAAT <- limAAT[order(-limAAT$PTM_Average_Annual_Trend), ]
limAAT$species <- factor(limAAT$species,
                         levels = unique(limAAT$species))
limAAT <- limAAT %>% 
  mutate(
    region = case_when(
      AAT_LowerCI < 0 & AAT_UpperCI < 0 ~ "decrease",
      AAT_LowerCI < 0 & AAT_UpperCI > 0 ~ "overlap",
      AAT_LowerCI > 0 & AAT_UpperCI > 0 ~ "increase"
    )
  )

# plotting

colors <- c(
  "decrease" <- "red",
  "increase" <- "blue",
  "overlap" <- "grey40"
)

aat2 <- ggplot(limAAT %>% filter(region != "overlap"),
               aes(x=species, y=PTM_Average_Annual_Trend,
                           color = region)) + 
  geom_point(size=3,
             alpha = 0.75) + 
  geom_segment(aes(x=species, 
                   xend=species, 
                   y=AAT_LowerCI,
                   yend=AAT_UpperCI),
               linewidth = 1.5) +
  geom_hline(yintercept = c(0), linetype="dotted") +
  ggtitle("Average Annual Trend per\nNCOS Species") + 
  xlab("Species") +
  ylab("Average Annual Trend") +
  coord_flip() + 
  theme_classic() + 
  scale_color_manual(values = colors) + 
  theme(aspect.ratio = 1.5,
        legend.position = "none",
        plot.title = element_text(hjust = 0.5,
                                  face = "bold",
                                  size = 24),
        axis.title.x = element_text(size = 20),
        axis.text.x = element_text(size = 16),
        axis.title.y = element_text(size = 20),
        axis.text.y = element_text(size = 16))
# trends match the "percent change in abundance" plots

ggsave(filename = paste0("CCBER_Species_AAT_no_grey_",
                         format(Sys.Date(), "%y-%m-%d"),
                         ".svg"),
       aat2,
       dpi = 1000,
       width = 9,
       height = 8,
       units = "in")

```
