---
title: "MV_work_2026-01-08"
output:
  html_document: default
  pdf_document: default
---

```{r setup,include = F}

library(here)
library(tidyverse)
library(popbayes)
library(R2jags)

datapath <- here("data","aggregated","MV_updated_aggregated.rds")
data <- readRDS(datapath)

```



```{r bootstrap data,include = F}

# formatting S x T matrix

year_abundance <- data %>% group_by(survey_year,species) %>% 
  summarise(abundance = sum(count))

community <- xtabs(abundance ~ survey_year + species,data = year_abundance)
community <- as.matrix(unclass(community))
community <- t(community)

```

```{r bootstrapping variables,include = F}

set.seed(234)

# calculating null model variables
# row 1 = m_1, for example

# m_i
null <- data.frame(m_i = 1:169)
for (i in 1:169){
  x = 0
  for (j in 1:8){
    x = x + community[i,j]
  }
  null$m_i[i] = x
}

# N
N <- 0
for (i in 1:169){
  N = N + null$m_i[i]
}
# 2696 individuals

# s_i
for (i in 1:169){
  null$s_i[i] = null$m_i[i] / N
}

null2 <- data.frame(n_j = 1:8,q_j = 1:8)
for (i in 1:8){
  null2$n_j[i] = sum(community[,i])
  null2$q_j[i] = null2$n_j[i] / N
}

# estimating undetected species

number_occurences <- data.frame(species_number_surveys_detected = 1:169)
for (i in 1:169){
  x = 0
  for (j in 1:8){
    if (community[i,j] > 0){
      x = x + 1
    }
  }
  number_occurences$species_number_surveys_detected[i] = x
}
q1_numbers <- which(number_occurences$species_number_surveys_detected == 1)
q1 <- length(q1_numbers)
q2_numbers <- which(number_occurences$species_number_surveys_detected == 2)
q2 <- length(q2_numbers)

S_bar <- (7 / 8) * ((q1 * (q1 - 1)) / (2 * (q2 + 1))) # 8.5, rounds to 9

# accounting for undetected species

undetected <- data.frame(m_i = 9 * NA,s_i = rep(0.5 * min(null$s_i),9))

null <- rbind(null,undetected) # adding 9 repetitions of 0.5 * minimum observed s_i

names <- append(rownames(community),paste0("undetected",1:9))

# testing simulation

null_test <- data.frame(sample_group = sample(1:8,N,prob = null2$q_j,replace = T),
                        sample_species = sample(names,N,prob = null$s_i,replace = T))

```

Simple linear model: $y_{ij} = \beta_{0i} + \beta_{1i}t_j + \epsilon_{ij}$ where $i = 1,\ 2,\ ...,\ S; S = \operatorname{number of species observed}; j = 1,\ 2,\ ...,\ T; T = \operatorname{number of sampling periods}$

$y_{ij}$: Simulated number of individuals of species $i$ in sample $j$.

$\beta_{0i}$: Regression intercept for species $i$.

$\beta_{1i}$: Regression slope for species $i$.

$\epsilon_{ij} \sim \mathcal{N}(0,\ \sigma_i^2)$: Error term

```{r observed regression plot,message = F,echo = F,fig.cap = "Observed trend lines for existing bird survey abundance data. Similar to the visualizations included in Gotelli's paper, species are trending upward and downward with a few heading into the negatives. Although the observed abundance scale is larger than those included in the paper, the replication of methods seems correct."}

observed <- data %>% group_by(survey_year,species) %>% 
  summarize(species_abundance = sum(count)) %>% 
  distinct(survey_year,species,species_abundance)



fig_observed <- ggplot(observed,aes(x = survey_year,y = species_abundance,color = species)) + 
  geom_smooth(method = "lm",se = F,fullrange = T) + theme_light() + 
  theme(legend.position = "none") +# ylim(c(0,50)) + 
  labs(x = "Year",y = "Observed Abundance",title = "Regression for Observed Data")
# if ylim is applied:
# stat_smooth removes 175 rows
# geom_smooth removes 183 rows

fig_observed
```

```{r simulated regression plot,message = F,echo = F,fig.cap = "Trend lines for one simulation of the null model. Compared to the abundance y-axis scale for the observed trendlines, the y-axis scale here is less by a factor of 20, thus the replication seems to be incorrect in some way."}

simulated <- null_test %>% group_by(sample_group,sample_species) %>% 
  tally %>% distinct(sample_group,sample_species,n)

ggplot(simulated,aes(x = sample_group,y = n,color = sample_species)) + 
  geom_smooth(method = "lm",se = F,fullrange = T) + theme_light() + 
  theme(legend.position = "none") +# ylim(c(0,50)) + 
  labs(x = "Year",y = "Simulated Abundance",title = "Regression for Simulated Null Moddl Data")
# if ylim is applied:
# geom_smooth removes 68 rows

```

```{r calcuating TCs,echo = F,fig.cap = "Calculated TC values for the observed data and the simulated null model data. While the simulated TC seems feasible, the observed TC of ~30,000 seems very off."}

# extracting beta1 coefficients for observed

species_observed <- unique(observed$species)
beta1_observed <- 1:length(species_observed)
for (i in 1:length(species_observed)){
  x <- lm(species_abundance ~ survey_year,
          data = observed[observed$species == species_observed[i],])
  beta1_observed[i] <- as.numeric(x$coefficients[2])
}
beta1_observed <- beta1_observed[!is.na(beta1_observed)]
# some coefficients are NA, caused by lack of data or consistent abundance

# calculating observed TC

intermediate1 <- 0
beta1bar_observed <- sum(beta1_observed)
for (i in 1:length(beta1_observed)){
  x <- (beta1_observed[i] - beta1bar_observed)^2
  intermediate1 = intermediate1 + x
}

TC_observed <- (1 / (length(species_observed) - 1)) * intermediate1
print(TC_observed)

# extractiing beta1 coefficients for simulated

species_simulated <- unique(simulated$sample_species)
beta1_simulated <- 1:length(species_simulated)
for (i in 1:length(species_simulated)){
  x <- lm(n ~ sample_group,
          data = simulated[simulated$sample_species == species_simulated[i],])
  beta1_simulated[i] <- as.numeric(x$coefficients[2])
}
beta1_simulated <- beta1_simulated[!is.na(beta1_simulated)]

# calculating simulated TC

intermediate2 <- 0
beta1bar_simulated <- sum(beta1_simulated)
for (i in 1:length(beta1_simulated)){
  x <- (beta1_simulated[i] - beta1bar_simulated)^2
  intermediate2 = intermediate2 + x
}

TC_simulated <- (1 / (length(species_simulated) - 1)) * intermediate2
print(TC_simulated)

# the observed TC is much larger than the simulated TC
# thus the observed data has much more heterogeneity

# observed TC is two orders of magnitude larger than any reported
# in gotelli paper - incorrect replication of methods/analysis?

```

Temporal change, $TC = \frac{1}{S - 1} \sum_{i = 1}^S(\beta_{1i} - \bar{\beta_1})^2$

Calculated TC values for the observed data and the simulated null model data. While the simulated TC seems feasible, the observed TC of ~30,000 seems very off.

