---
title: "MV_2025-11-19"
output:
  pdf_document: default
  html_document: default
---

todo

make aggregated include all other variables such as water level etc
also save aggregated as an .rds to save data types

google doc of stuff to do long term?



done

facet wrap substrate line graph to separate substrates - higher proportion vs lower proportions or logical grouping (water, herbs)

make NMDS plots not have giant triangle - single, smaller point for differentiability

add species to seasons, maximum nmds

sequential color palette or arrows for maximum nmds instead of current color grouping system



```{r,include = F}

library(here)
library(tidyverse)
library(patchwork)
library(vegan)
library(ggrepel)

MVaggregatedpath <- here("data","aggregated","MV_aggregated.csv")
aggregated <- read_csv(MVaggregatedpath)

```

```{r substrate observation proportions per survey year patchwork facet,out.width = "100%",fig.align = "center"}

### substrate groupings for visualization differentiation
### based on similar observation proportions and logic
# - vegetation
# - air + bare ground + unknown
# - water + mud
# - man-made: bridge, building, fence
#  + other
# - habitat features

relative_proportions <- aggregated %>% group_by(Survey_Year,Substrate) %>%
  summarize(substrate_count = sum(Count)) %>% 
  mutate(total = sum(substrate_count)) %>% 
  mutate(proportion = substrate_count / total,grouping = case_when(
    grepl("Vegetation",Substrate) ~ 1,
    Substrate == "Air" ~ 2,
    Substrate == "Bare ground" ~ 2,
    Substrate == "Unknown" ~ 2,
    grepl("Water",Substrate) ~ 3,
    Substrate == "Mud" ~ 3,
    Substrate == "Bridge" ~ 4,
    Substrate == "Building" ~ 4,
    Substrate == "Fence" ~ 4,
    Substrate == "Other" ~ 4,
    grepl("Habitat",Substrate) ~ 5))
# grepl returns boolean vector for string component match

chosen <- relative_proportions %>% 
  filter(Substrate != "Air/Flyover" & Substrate != "Flyover" & Substrate != "Rock" & 
           Substrate != "telephone wire" & Substrate != "Vernal Pool")
# removes 12 rows, 8 flyover rows and 1 of each of the others



line1 <- ggplot(chosen %>% filter(grouping == 1),
                aes(x = Survey_Year,y = proportion,group = Substrate,color = Substrate)) + 
  geom_line(alpha = 0.85,linewidth = 1) + theme_light() + 
  facet_wrap(~ grouping,scales = "free_y",ncol = 1,dir = "v") + 
  labs(x = "Survey Year",y = "Proportion") + 
  scale_x_continuous(breaks = seq(1,8,by = 1))

line2 <- ggplot(chosen %>% filter(grouping == 2),
                aes(x = Survey_Year,y = proportion,group = Substrate,color = Substrate)) + 
  geom_line(alpha = 0.85,linewidth = 1) + theme_light() + 
  facet_wrap(~ grouping,scales = "free_y",ncol = 1,dir = "v") + 
  labs(x = "Survey Year",y = "Proportion") + 
  scale_x_continuous(breaks = seq(1,8,by = 1))

line3 <- ggplot(chosen %>% filter(grouping == 3),
                aes(x = Survey_Year,y = proportion,group = Substrate,color = Substrate)) + 
  geom_line(alpha = 0.85,linewidth = 1) + theme_light() + 
  facet_wrap(~ grouping,scales = "free_y",ncol = 1,dir = "v") + 
  labs(x = "Survey Year",y = "Proportion") + 
  scale_x_continuous(breaks = seq(1,8,by = 1))

line4 <- ggplot(chosen %>% filter(grouping == 4),
                aes(x = Survey_Year,y = proportion,group = Substrate,color = Substrate)) + 
  geom_line(alpha = 0.85,linewidth = 1) + theme_light() + 
  facet_wrap(~ grouping,scales = "free_y",ncol = 1,dir = "v") + 
  labs(x = "Survey Year",y = "Proportion") + 
  scale_x_continuous(breaks = seq(1,8,by = 1))

line5 <- ggplot(chosen %>% filter(grouping == 5),
                aes(x = Survey_Year,y = proportion,group = Substrate,color = Substrate)) + 
  geom_line(alpha = 0.85,linewidth = 1) + theme_light() + 
  facet_wrap(~ grouping,scales = "free_y",ncol = 1,dir = "v") + 
  labs(x = "Survey Year",y = "Proportion") + 
  scale_x_continuous(breaks = seq(1,8,by = 1))



plot1 <- (line1 + line2) / (line3 + line4) / (line5 + plot_spacer()) + 
  plot_annotation(title = "Substrate Relative Observation Proportion per Survey Year",
                  theme = theme(plot.title = element_text(hjust = 0.5)))
ggsave(paste0("Substrate_Proportions_1_",format(Sys.Date(),"%y-%m-%d"),".pdf"),
       plot1,width = 200,height = 180,units = "mm",)

```



```{r seasons nmds}

community1 <- xtabs(Count ~ Observation_Date + Species,data = aggregated)
community1 <- as.matrix(unclass(community1))
nmds1 <- metaMDS(community1,autotransform = F,distance = "bray",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

season <- aggregated %>% distinct(Observation_Date,Season)
season$Observation_Date <- as.character(season$Observation_Date)
top_species_pseason <- aggregated %>% group_by(Season,Species) %>%
  summarise(total = sum(Count)) %>% slice_max(order_by = total,n = 5)

nmds1_site_scores <- as.data.frame(scores(nmds1)$sites) %>% 
  rownames_to_column(var = "Observation_Date") %>% 
  left_join(season)
nmds1_species_scores <- as.data.frame(scores(nmds1,"species"))
nmds1_species_scores$Species <- rownames(nmds1_species_scores)
nmds1_species_scores_joined <- right_join(nmds1_species_scores,top_species_pseason)

season_centroid <- nmds1_site_scores %>% group_by(Season) %>% summarise(
  axis1 = mean(NMDS1),
  axis2 = mean(NMDS2)) %>% ungroup()
season_shape <- nmds1_site_scores %>% group_by(Season) %>% slice(chull(NMDS1,NMDS2))

nmdsfigure1 <- ggplot() + geom_point(data = nmds1_site_scores,aes(x = NMDS1,y = NMDS2,color = Season),
                      size = 1.5) + 
  geom_point(data = season_centroid,aes(x = axis1,y = axis2,color = Season),
             size = 6,shape = 17) + 
  geom_polygon(data = season_shape,aes(x = NMDS1,y = NMDS2,fill = Season,group = Season),
               alpha = 0.25) +
  geom_text_repel(data = nmds1_species_scores_joined,
                  aes(x = NMDS1,y = NMDS2,label = Species),max.overlaps = 500,size = 3) + 
  annotate("text",x = 0,y = 1.25,label = paste("2D stress =",round(nmds1$stress,3))) + 
  theme_light() + xlim(c(-1,1)) + ylim(c(-1.5,1.5)) +
  labs(title = "Species Abundance by Season NMDS") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave(paste0("Season_NMDS_",format(Sys.Date(),"%y-%m-%d"),".pdf"),nmdsfigure1,
       width = 200,units = "mm")

```



```{r maximum survey count per survey year nmds}

year_max <- aggregated %>% group_by(Species,Survey_Year) %>% 
  summarise(max_count = max(Count))
top_species_pyear <- year_max %>% group_by(Survey_Year) %>% 
  slice_max(order_by = max_count,n = 3)

community2 <- xtabs(max_count ~ Survey_Year + Species,data = year_max)
community2 <- as.matrix(unclass(community2))

nmds2 <- metaMDS(community2,autotransform = F,distance = "bray",engine = "monoMDS",
                 k = 2,weakties = T,model = "global",maxit = 300,try = 40,
                 trymax = 100,trace = F)

nmds2_site_scores <- as.data.frame(scores(nmds2)$sites) %>% 
  rownames_to_column(var = "Survey_Year")
nmds2_site_scores$Survey_Year <- as.numeric(nmds2_site_scores$Survey_Year)
nmds2_site_scores <- nmds2_site_scores %>% arrange(Survey_Year) %>% 
  mutate(xend = lead(NMDS1),yend = lead(NMDS2))
nmds2_species_scores <- as.data.frame(scores(nmds2,"species"))
nmds2_species_scores$Species <- rownames(nmds2_species_scores)
nmds2_species_scores_joined <- right_join(nmds2_species_scores,top_species_pyear)
nmds2_species_scores_joined$Survey_Year <- as.numeric(nmds2_species_scores_joined$Survey_Year)

nmdsfigure2 <- ggplot() + geom_point(data = nmds2_site_scores,
                      aes(x = NMDS1,y = NMDS2,color = Survey_Year),size = 3) + 
  geom_segment(data = nmds2_site_scores,
               aes(x = NMDS1,y = NMDS2,xend = xend,yend = yend,color = Survey_Year),
                   arrow = arrow(length = unit(0.25,"cm"))) + 
  geom_text_repel(data = nmds2_species_scores_joined,
                  aes(x = NMDS1,y = NMDS2,label = Species),max.overlaps = 500,size = 3) + 
  annotate("text",x = -0.2,y = 0.1,label = paste("2D stress =",round(nmds2$stress,3))) + 
  scale_color_viridis_c(option = "C",alpha = 0.8) + 
  theme_light() + labs(title = "Maximum Species Abundance by Survey per Year NMDS") + 
  theme(plot.title = element_text(hjust = 0.5))

ggsave(paste0("Maximum_Survey_per_Year_NMDS_",format(Sys.Date(),"%y-%m-%d"),".pdf"),nmdsfigure2,
       width = 200,units = "mm")

```


